Filippo Gastaldello - 17/03/2026

```{r}
library(tidyverse)
library(UpSetR)
```

# Paths
```{r}
# Inputs
gigatable_path <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/gigatable.csv"

# Plots output
out_dir <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/methods_comparison/"
```

```{r}
gigatable <- read_csv(gigatable_path)
```

# Thresholds
```{r}
FDR_padj_thresh         <- 0.05
log2FC_lmm_phase_thresh <- 1
```

# LMM-GSEA-ORA Intersection

Check overlap in the lists of deregulated programs according to LMM, GSEA and ORA
```{r}
upset_data <- list(
                LMM = gigatable %>% 
                        filter(FDR_lmm_phase<0.05,
                               abs(log2FC_lmm_phase)>1) %>% 
                        mutate(label = paste0(cell_type, "_", program_short)) %>% 
                        pull(label) %>% 
                        unique(),
                
                GSEA_sc = gigatable %>% 
                                filter(!is.na(padj_GSEA_sc)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                
                GSEA_pb = gigatable %>% 
                                filter(!is.na(padj_GSEA_pb)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                
                ORA_sc = gigatable %>% 
                                filter(!is.na(padj_ORA_UP_sc) | !is.na(padj_ORA_DOWN_sc)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                
                ORA_pb = gigatable %>% 
                                filter(!is.na(padj_ORA_UP_pb) | !is.na(padj_ORA_DOWN_pb)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                
                ORA_both = gigatable %>% 
                                filter(!is.na(padj_ORA_UP_both) | !is.na(padj_ORA_DOWN_both)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique()
)
# RUN THIS PART IN CONSOLE
pdf(file = paste0(out_dir, "upset_LMM_GSEA_ORA.pdf"), width = 9, height = 5, onefile = FALSE)
upset(fromList(upset_data),
      order.by = "freq",
      nsets = 6)
dev.off()
```

# LMM - GSEA
## LMM to GSEA
Volcano plot of program activation (SPECTRA scores) according to LMM, with information about significance in GSEA analysis.
```{r}
plot_data <- gigatable %>%  
                  filter(!is.na(log2FC_lmm_phase)) %>% 
                  dplyr::select(cell_type,
                                program,
                                program_short,
                                FDR_lmm_phase,
                                log2FC_lmm_phase,
                                padj_GSEA_sc,
                                NES_GSEA_sc,
                                padj_GSEA_pb,
                                NES_GSEA_pb) %>% 
                  mutate(also_sig_in = case_when(
                                                !is.na(padj_GSEA_sc) & !is.na(padj_GSEA_pb) ~ "both",
                                                !is.na(padj_GSEA_sc)                        ~ "sc",
                                                !is.na(padj_GSEA_pb)                        ~ "pb",
                                                TRUE                                        ~ NA_character_
                                                ),
                         also_sig_in = as.factor(also_sig_in),
                         diff_activated = case_when(
                                                FDR_lmm_phase < FDR_padj_thresh & log2FC_lmm_phase > log2FC_lmm_phase_thresh ~ "UP",
                                                FDR_lmm_phase < FDR_padj_thresh & log2FC_lmm_phase < -log2FC_lmm_phase_thresh ~ "DOWN",
                                                TRUE ~ "NO"
                                                     )
                         )

p <- plot_data %>% ggplot(aes(x = log2FC_lmm_phase, y = -log10(FDR_lmm_phase), color = diff_activated, shape = also_sig_in)) + 
                geom_point(size = 2) +
                geom_vline(
                        xintercept = c(-log2FC_lmm_phase_thresh, log2FC_lmm_phase_thresh),
                        col = 'black',
                        linetype = 'dashed'
                        ) +
                geom_hline(
                        yintercept = -log10(FDR_padj_thresh),
                        col = 'black',
                        linetype = 'dashed'
                        ) +
                scale_colour_manual(
                        values = c("UP" = "red", "DOWN" = "purple", "NO" = "grey")
                        ) +
                scale_shape_manual(
                        values   = c("sc" = 15, "pb" = 17, "both" = 18),
                        na.value = 16,
                        breaks   = c("sc", "pb", "both"),
                        labels   = c("Single cell GSEA", "Pseudobulk GSEA", "Both")
                        ) +
                xlim(
                        (plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN") %>% pull(log2FC_lmm_phase) %>% min()) * 1.1,
                        (plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN") %>% pull(log2FC_lmm_phase) %>% max()) * 1.1
                        ) +
                ylim(
                        0,
                        (plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN") %>% pull(FDR_lmm_phase) %>% log10() %>% min()) * -1.1
                        ) +
                xlab("log2(Activation FC)") +
                ylab("-log10(FDR)") +
                labs(
                        title    = "Program deregulation according to LMM analyzing Spectra scores.",
                        subtitle = "Between parentesis is the cell type in which such problem was found deregulated.",
                        color    = "Program deregulation",
                        shape    = "Also found in"
                        ) +
                geom_label_repel(
                        data               = plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN"),
                        aes(label = paste0("(", cell_type, ")", " ", program_short)),
                        size               = 2.5,           
                        max.overlaps       = Inf,   
                        box.padding        = 0.5,
                        point.padding      = 0.2,
                        min.segment.length = 0,
                        nudge_x            = .15,
                        nudge_y            = .5
                        ) +
                theme_minimal()

ggsave(plot = p,
       filename = paste0(out_dir, "LMM-GSEA/LMM_volcano_GSEA.pdf"),
       device = "pdf",
       width = 11,
       height = 8,
       create.dir = TRUE)
write_csv(plot_data, paste0(out_dir, "LMM-GSEA/LMM_volcano_GSEA_plot_data.csv"))
```

## GSEA to LMM
Given programs enriched in GSEA analysis, check if they are significantly altered according to LMM analysis of spectra scores, or if at least they have concordant sign
```{r}
configs <- list(
                sc = list(
                        padj  = "padj_GSEA_sc",
                        NES   = "NES_GSEA_sc",
                        title = "single cell"
                ),
                pb = list(
                        padj  = "padj_GSEA_pb",
                        NES   = "NES_GSEA_pb",
                        title = "pseudobulk"
                )
)

plots <- lapply(names(configs), function(name){
                
                conf <- configs[[name]]
                
                plot_data <- gigatable %>%
                                filter(!is.na(.data[[conf$padj]])) %>% 
                                dplyr::select(cell_type,
                                              program_short,
                                              FDR_lmm_phase,
                                              log2FC_lmm_phase,
                                              conf$padj,
                                              conf$NES) %>% 
                                mutate(concordant = case_when(
                                                        sign(log2FC_lmm_phase) == sign(.data[[conf$NES]]) & sign(log2FC_lmm_phase) == -1 ~ "DOWN",
                                                        sign(log2FC_lmm_phase) == sign(.data[[conf$NES]]) & sign(log2FC_lmm_phase) == 1  ~ "UP",
                                                        TRUE                                                                             ~ "no"
                                                        ),
                                       LMM_sig    = ifelse(FDR_lmm_phase < FDR_padj_thresh, "yes", "no")
                                       )
                
                p <- plot_data %>%
                        ggplot(aes(x = .data[[conf$NES]], y = log2FC_lmm_phase, color = concordant, shape = LMM_sig)) +
                                geom_point(size = 2) +
                                scale_color_manual(
                                        values  = c("UP" = "red", "DOWN" = "purple", "no" = "grey"),
                                        breaks  = c("UP", "DOWN", "no"),
                                        labels  = c("Both up", "Both down", "Discordant")
                                        ) +
                                scale_shape_manual(
                                        values = c("yes" = 16, "no" = 4),
                                        breaks = c("yes", "no")
                                        ) +
                                geom_vline(
                                        xintercept = 0,
                                        col        = "black",
                                        linetype   = "dashed"
                                        ) +
                                geom_hline(
                                        yintercept = 0,
                                        col        = "black",
                                        linetype   = "dashed"
                                        ) +
                                labs(
                                        title = paste0("Enriched programs in GSEA analysis according to ", conf$title," DEGs "),
                                        color = "GSEA NES vs LMM FC sign",
                                        shape = "Significant (FDR<0.05) also in \nLMM analysis of Spectra scores"
                                        ) +
                                xlab("Pseudobulk GSEA NES") +
                                ylab("LMM log2(Activation FC)") +
                                geom_label_repel(
                                        data               = plot_data %>% filter(!concordant == "no"),
                                        aes(label = paste0("(", cell_type, ")", " ", program_short)),
                                        size               = 2.5,           
                                        max.overlaps       = Inf,   
                                        box.padding        = 0.5,
                                        point.padding      = 0.2,
                                        min.segment.length = 0,
                                        nudge_x            = .15,
                                        nudge_y            = .5
                                        ) +
                                theme_minimal()
                
                ggsave(plot = p,
                       filename = paste0(out_dir, "LMM-GSEA/GSEA_", name, "_scatter_LMM.pdf"),
                       device = "pdf",
                       width = 13,
                       height = 10,
                       create.dir = TRUE)
                write_csv(plot_data, paste0(out_dir, "LMM-GSEA/GSEA_", name, "_scatter_LMM_plot_data.csv"))
})



```

# LMM-ORA
## LMM to ORA

Volcano plot of program activation (SPECTRA scores) according to LMM, with information about significance in ORA analysis.
```{r}
plot_data <- gigatable %>% 
                filter(FDR_lmm_phase <- 0.05) %>% 
                select(cell_type,
                       program_short)
```

