Filippo Gastaldello - 17/03/2026

```{r}
library(tidyverse)
library(ComplexHeatmap)
library(ggrepel)
```

# Paths
```{r}
# Inputs
gigatable_path <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/gigatable.csv"

# Plots output
out_dir <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/methods_comparison/"
```

```{r}
gigatable <- read_csv(gigatable_path)
```

# Thresholds
```{r}
FDR_padj_thresh         <- 0.05
log2FC_lmm_phase_thresh <- 1
```

# LMM-GSEA-ORA

## Intersection (UpSet plot)

Check overlap in the lists of deregulated programs according to LMM, GSEA and ORA

```{r}
upset_data <- list(
  LMM = gigatable %>% 
    filter(FDR_lmm_phase<0.05,
           abs(log2FC_lmm_phase)>1) %>% 
    mutate(label = paste0(cell_type, "_", program_short)) %>% 
    pull(label) %>% 
    unique(),
  
  GSEA_sc = gigatable %>% 
    filter(!is.na(padj_GSEA_sc)) %>% 
    mutate(label = paste0(cell_type, "_", program_short)) %>% 
    pull(label) %>% 
    unique(),
  
  GSEA_pb = gigatable %>% 
    filter(!is.na(padj_GSEA_pb)) %>% 
    mutate(label = paste0(cell_type, "_", program_short)) %>% 
    pull(label) %>% 
    unique(),
  
  ORA_sc = gigatable %>% 
    filter(!is.na(padj_ORA_UP_sc) | !is.na(padj_ORA_DOWN_sc)) %>% 
    mutate(label = paste0(cell_type, "_", program_short)) %>% 
    pull(label) %>% 
    unique(),
  
  ORA_pb = gigatable %>% 
    filter(!is.na(padj_ORA_UP_pb) | !is.na(padj_ORA_DOWN_pb)) %>% 
    mutate(label = paste0(cell_type, "_", program_short)) %>% 
    pull(label) %>% 
    unique(),
  
  ORA_both = gigatable %>% 
    filter(!is.na(padj_ORA_UP_both) | !is.na(padj_ORA_DOWN_both)) %>% 
    mutate(label = paste0(cell_type, "_", program_short)) %>% 
    pull(label) %>% 
    unique()
)
```

### Horizontal
```{r}
upset_data <- list(
                LMM = gigatable %>% 
                        filter(FDR_lmm_phase<0.05,
                               abs(log2FC_lmm_phase)>1) %>% 
                        mutate(label = paste0(cell_type, "_", program_short)) %>% 
                        pull(label) %>% 
                        unique(),
                GSEA_sc = gigatable %>% 
                                filter(!is.na(padj_GSEA_sc)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                GSEA_pb = gigatable %>% 
                                filter(!is.na(padj_GSEA_pb)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                ORA_sc = gigatable %>% 
                                filter(!is.na(padj_ORA_UP_sc) | !is.na(padj_ORA_DOWN_sc)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                ORA_pb = gigatable %>% 
                                filter(!is.na(padj_ORA_UP_pb) | !is.na(padj_ORA_DOWN_pb)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique(),
                ORA_both = gigatable %>% 
                                filter(!is.na(padj_ORA_UP_both) | !is.na(padj_ORA_DOWN_both)) %>% 
                                mutate(label = paste0(cell_type, "_", program_short)) %>% 
                                pull(label) %>% 
                                unique()

)

# RUN THIS PART IN CONSOLE
pdf(file = paste0(out_dir, "upset_LMM_GSEA_ORA.pdf"), width = 9, height = 5, onefile = FALSE)
upset(fromList(upset_data),
      order.by = "freq",
      nsets = 6)
dev.off()
```

```{r}
m <- make_comb_mat(upset_data)
m <- m[comb_degree(m) > 1] 

m_transposed <- t(m)

all_codes <- comb_name(m_transposed)
all_sizes <- comb_size(m_transposed)
all_degrees <- comb_degree(m_transposed) 

lmm_index <- which(set_name(m_transposed) == "LMM")

plot_order <- order(all_degrees, all_sizes, decreasing = TRUE)

all_labels <- sapply(seq_along(all_codes), function(i) {
  
  includes_lmm <- substr(all_codes[i], lmm_index, lmm_index) == "1"
  
  if ((all_degrees[i] >= 3 && all_sizes[i] < 7) || includes_lmm) {
    items <- extract_comb(m, all_codes[i])
    return(paste(items, collapse = "\n"))
  } else {
    return("") 
  }
})

combined_right_annotation <- rowAnnotation(
  "Size" = anno_text(
    as.character(all_sizes), 
    just = "center", 
    gp = gpar(fontsize = 10, fontface = "bold")
  ),
  "Items" = anno_text(
    all_labels, 
    just = "left", 
    location = unit(0, "npc"), 
    gp = gpar(fontsize = 8)
  ),
  gap = unit(4, "mm") 
)

pdf(file = paste0(out_dir, "upset_LMM_GSEA_ORA_vertical.pdf"), width = 11, height = 20, onefile = FALSE)

draw(UpSet(
  m_transposed, 
  comb_order = plot_order,
  right_annotation = combined_right_annotation,
  column_names_side = "top",
  column_names_rot  = 45
))

dev.off()
```

# LMM - GSEA
## LMM to GSEA
Volcano plot of program activation (SPECTRA scores) according to LMM, with information about significance in GSEA analysis.
```{r}
plot_data <- gigatable %>%  
                  filter(!is.na(log2FC_lmm_phase)) %>% 
                  dplyr::select(cell_type,
                                program,
                                program_short,
                                FDR_lmm_phase,
                                log2FC_lmm_phase,
                                padj_GSEA_sc,
                                NES_GSEA_sc,
                                padj_GSEA_pb,
                                NES_GSEA_pb) %>% 
                  mutate(also_sig_in = case_when(
                                                !is.na(padj_GSEA_sc) & !is.na(padj_GSEA_pb) ~ "both",
                                                !is.na(padj_GSEA_sc)                        ~ "sc",
                                                !is.na(padj_GSEA_pb)                        ~ "pb",
                                                TRUE                                        ~ NA_character_
                                                ),
                         also_sig_in = as.factor(also_sig_in),
                         diff_activated = case_when(
                                                FDR_lmm_phase < FDR_padj_thresh & log2FC_lmm_phase > log2FC_lmm_phase_thresh ~ "UP",
                                                FDR_lmm_phase < FDR_padj_thresh & log2FC_lmm_phase < -log2FC_lmm_phase_thresh ~ "DOWN",
                                                TRUE ~ "NO"
                                                     )
                         )

p <- plot_data %>% ggplot(aes(x = log2FC_lmm_phase, y = -log10(FDR_lmm_phase), color = diff_activated, shape = also_sig_in)) + 
                geom_point(size = 2) +
                geom_vline(
                        xintercept = c(-log2FC_lmm_phase_thresh, log2FC_lmm_phase_thresh),
                        col = 'black',
                        linetype = 'dashed'
                        ) +
                geom_hline(
                        yintercept = -log10(FDR_padj_thresh),
                        col = 'black',
                        linetype = 'dashed'
                        ) +
                scale_colour_manual(
                        values = c("UP" = "red", "DOWN" = "purple", "NO" = "grey")
                        ) +
                scale_shape_manual(
                        values   = c("sc" = 15, "pb" = 17, "both" = 18),
                        na.value = 16,
                        breaks   = c("sc", "pb", "both"),
                        labels   = c("Single cell GSEA", "Pseudobulk GSEA", "Both")
                        ) +
                xlim(
                        (plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN") %>% pull(log2FC_lmm_phase) %>% min()) * 1.1,
                        (plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN") %>% pull(log2FC_lmm_phase) %>% max()) * 1.1
                        ) +
                ylim(
                        0,
                        (plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN") %>% pull(FDR_lmm_phase) %>% log10() %>% min()) * -1.1
                        ) +
                xlab("log2(Activation FC)") +
                ylab("-log10(FDR)") +
                labs(
                        title    = "Program deregulation according to LMM analyzing Spectra scores.",
                        subtitle = "Between parentesis is the cell type in which such problem was found deregulated.",
                        color    = "Program deregulation",
                        shape    = "Also found in"
                        ) +
                geom_label_repel(
                        data               = plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN"),
                        aes(label = paste0("(", cell_type, ")", " ", program_short)),
                        size               = 2.5,           
                        max.overlaps       = Inf,   
                        box.padding        = 0.5,
                        point.padding      = 0.2,
                        min.segment.length = 0,
                        nudge_x            = .15,
                        nudge_y            = .5
                        ) +
                theme_minimal()

ggsave(plot = p,
       filename = paste0(out_dir, "LMM-GSEA/LMM_volcano_GSEA.pdf"),
       device = "pdf",
       width = 11,
       height = 8,
       create.dir = TRUE)
write_csv(plot_data, paste0(out_dir, "LMM-GSEA/LMM_volcano_GSEA_plot_data.csv"))
```

## GSEA to LMM
Given programs enriched in GSEA analysis, check if they are significantly altered according to LMM analysis of spectra scores, or if at least they have concordant sign
```{r}
configs <- list(
                sc = list(
                        padj  = "padj_GSEA_sc",
                        NES   = "NES_GSEA_sc",
                        title = "single cell"
                ),
                pb = list(
                        padj  = "padj_GSEA_pb",
                        NES   = "NES_GSEA_pb",
                        title = "pseudobulk"
                )
)

lapply(names(configs), function(name){
                
        conf <- configs[[name]]
        
        plot_data <- gigatable %>%
                        filter(!is.na(.data[[conf$padj]])) %>% 
                        dplyr::select(cell_type,
                                      program_short,
                                      FDR_lmm_phase,
                                      log2FC_lmm_phase,
                                      conf$padj,
                                      conf$NES) %>% 
                        mutate(concordant = case_when(
                                                sign(log2FC_lmm_phase) == sign(.data[[conf$NES]]) & sign(log2FC_lmm_phase) == -1 ~ "DOWN",
                                                sign(log2FC_lmm_phase) == sign(.data[[conf$NES]]) & sign(log2FC_lmm_phase) == 1  ~ "UP",
                                                TRUE                                                                             ~ "no"
                                                ),
                               LMM_sig    = ifelse(FDR_lmm_phase < FDR_padj_thresh, "yes", "no")
                               )
        
        p <- plot_data %>%
                ggplot(aes(x = .data[[conf$NES]], y = log2FC_lmm_phase, color = concordant, shape = LMM_sig)) +
                        geom_point(size = 2) +
                        scale_color_manual(
                                values  = c("UP" = "red", "DOWN" = "purple", "no" = "grey"),
                                breaks  = c("UP", "DOWN", "no"),
                                labels  = c("Both up", "Both down", "Discordant")
                                ) +
                        scale_shape_manual(
                                values = c("yes" = 16, "no" = 4),
                                breaks = c("yes", "no")
                                ) +
                        geom_vline(
                                xintercept = 0,
                                col        = "black",
                                linetype   = "dashed"
                                ) +
                        geom_hline(
                                yintercept = 0,
                                col        = "black",
                                linetype   = "dashed"
                                ) +
                        labs(
                                title = paste0("Enriched programs in GSEA analysis according to ", conf$title," DEGs"),
                                color = "GSEA NES vs LMM FC sign",
                                shape = "Significant (FDR<0.05) also in \nLMM analysis of Spectra scores"
                                ) +
                        xlab("Pseudobulk GSEA NES") +
                        ylab("LMM log2(Activation FC)") +
                        geom_label_repel(
                                data               = plot_data %>% filter(!concordant == "no"),
                                aes(label = paste0("(", cell_type, ")", " ", program_short)),
                                size               = 2.5,           
                                max.overlaps       = Inf,   
                                box.padding        = 0.5,
                                point.padding      = 0.2,
                                min.segment.length = 0,
                                nudge_x            = 0.15,
                                nudge_y            = 0.5
                                ) +
                        theme_minimal()
        
        ggsave(plot       = p,
               filename   = paste0(out_dir, "LMM-GSEA/GSEA_", name, "_scatter_LMM.pdf"),
               device     = "pdf",
               width      = 13,
               height     = 10,
               create.dir = TRUE)
        write_csv(plot_data, paste0(out_dir, "LMM-GSEA/GSEA_", name, "_scatter_LMM_plot_data.csv"))
})
```

# LMM-ORA
## ORA to LMM

Given programs enriched in ORA analysis, check if they are significantly altered according to LMM analysis of spectra scores, or if at least they have concordant sign
```{r}
configs <- list(
                sc = list(
                        padj_up        = "padj_ORA_UP_sc",
                        padj_down      = "padj_ORA_DOWN_sc",
                        generatio_up   = "GeneRatio_ORA_UP_sc",
                        generatio_down = "GeneRatio_ORA_DOWN_sc",
                        title          = "single cell"
                ),
                pb = list(
                        padj_up        = "padj_ORA_UP_pb",
                        padj_down      = "padj_ORA_DOWN_pb",
                        generatio_up   = "GeneRatio_ORA_UP_pb",
                        generatio_down = "GeneRatio_ORA_DOWN_pb",
                        title          = "pseudobulk"
                ),
                both = list(
                        padj_up        = "padj_ORA_UP_both",
                        padj_down      = "padj_ORA_DOWN_both",
                        generatio_up   = "GeneRatio_ORA_UP_both",
                        generatio_down = "GeneRatio_ORA_DOWN_both",
                        title          = "both single cell and pseudobulk"
                )
)

lapply(names(configs), function(name){

        conf <- configs[[name]]

        plot_data <- gigatable %>% 
                        filter(!is.na(.data[[conf$padj_up]]) | !is.na(.data[[conf$padj_down]])) %>% 
                        filter(!is.na(log2FC_lmm_phase)) %>% 
                        dplyr::select(cell_type,
                                      program_short,
                                      FDR_lmm_phase,
                                      log2FC_lmm_phase,
                                      conf$padj_down,
                                      conf$generatio_down,
                                      conf$padj_up,
                                      conf$generatio_up) %>% 
                        mutate(!!conf$generatio_down := .data[[conf$generatio_down]] * -1) %>%  # Flip sign of gene ratio for down regulated programs to plot in negative quadrant
                                                                                                # '!!' evaluate the content of the variable 'conf$generatio_down' and use it as columns name
                                                                                                # ':=' allows the assignment to a dynamic evaluation
                        mutate(up_concordant   = ifelse(!is.na(.data[[conf$padj_up]]) & sign(log2FC_lmm_phase) == 1, "yes","no"),
                               up_concordant   = ifelse(is.na(up_concordant), "no", up_concordant),
                               down_concordant = ifelse(!is.na(.data[[conf$padj_down]]) & sign(log2FC_lmm_phase) == -1, "yes", "no"),
                               down_concordant = ifelse(is.na(down_concordant), "no", down_concordant),
                               lmm_sig         = case_when(
                                                        FDR_lmm_phase < FDR_padj_thresh    ~ "yes",
                                                        TRUE                    ~ "no"
                               ))
        # Transform the data first
        plot_data_long <- plot_data %>%
          # Combine UP and DOWN generatio into one X-axis column
          pivot_longer(cols = c(conf$generatio_up, conf$generatio_down),
                       names_to = "ora_direction",
                       values_to = "generatio") %>%
          mutate(concordance_status = case_when(
                  ora_direction == conf$generatio_up & up_concordant == "yes" ~ "UP_yes",
                  ora_direction == conf$generatio_down & down_concordant == "yes" ~ "DOWN_yes",
                  TRUE ~ "no"))

        # Plot with a single geom_point and a single color scale
        p <- plot_data_long %>% ggplot(aes(x = generatio, y = log2FC_lmm_phase, shape = lmm_sig, color = concordance_status)) +
                                geom_point() +
                                scale_shape_manual(
                                        values = c("yes" = 16, "no" = 4)
                                        ) +
                                scale_color_manual(
                                        name   = "Concordance",
                                        values = c("UP_yes" = "red", "DOWN_yes" = "purple", "no" = "grey"),
                                        labels = c("UP_yes" = "UP Concordant", "DOWN_yes" = "DOWN Concordant", "no" = "Not Concordant")
                                        ) +
                                geom_hline(
                                        yintercept = 0,
                                        col = "black",
                                        linetype = "dashed"
                                        ) +
                                geom_vline(
                                        xintercept = 0,
                                        col = "black",
                                        linetype = "dashed"
                                        ) +
                                labs(
                                        title = paste0("Enriched programs in ORA analysis according to ", conf$title," DEGs"),
                                        shape = "Also significant (FDR < 0.05)\nin LMM analysis"
                                        ) +
                                xlab("ORA gene ratio") +
                                ylab("LMM log2(Activation FC)") +
                                geom_label_repel(
                                        data               = plot_data_long %>% filter(!concordance_status == "no" & lmm_sig == "yes"),
                                        aes(label = paste0("(", cell_type, ")", " ", program_short)),
                                        size               = 2.5,           
                                        max.overlaps       = Inf,   
                                        box.padding        = 0.5,
                                        point.padding      = 0.2,
                                        min.segment.length = 0,
                                        nudge_x            = 0.15,
                                        nudge_y            = 0.5
                                        ) +
                                theme_minimal()
        ggsave(plot = p,
               filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/methods_comparison/LMM-ORA/ORA_", name, "_scatter_LMM.pdf"),
               width = 10,
               height = 7,
               device = "pdf",
               create.dir = TRUE)
})
```

# BIOLOGICAL RELEVANCE

Compute distributions of relevance scores (final_score) from the MAGIC REPORT (cytopus_ipf_full_program_scores.json) between deregulations detected with SPECTRA vs DEA.
```{r}
# Load program scores, used to compute biological relevance
program_scores_IPF <- read_tsv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/resources/cytopus_relevance_IPF/cytopus_ipf/cytopus_ipf_full_program_scores.tsv")

DEA_cols  <- colnames(gigatable)[grepl("GSEA", colnames(gigatable)) | grepl("ORA", colnames(gigatable))]
to_select <- c("cell_type", "program_short")

# Extract celltype-program deregulations detected by SPECTRA, DEA abd both
programs_spectra <- gigatable %>% 
                        filter(FDR_lmm_phase < FDR_padj_thresh, abs(log2FC_lmm_phase) > log2FC_lmm_phase_thresh,            # Significant in LMM
                               if_all(DEA_cols, is.na)) %>%                                                                 # Not significant in any DEA method
                        dplyr::select(all_of(to_select)) %>% 
                        unique() %>% 
                        left_join(program_scores_IPF %>%
                                        dplyr::select(program_name,
                                                      final_score),
                                  by = join_by("program_short" == "program_name")) %>% 
                        mutate(analysis = "Spectra")
programs_dea     <- gigatable %>% 
                        filter(!if_all(DEA_cols, is.na),                                                                    # Significant in at least one DEA method
                               !(FDR_lmm_phase < FDR_padj_thresh & abs(log2FC_lmm_phase) > log2FC_lmm_phase_thresh)) %>%    # Not significant in LMM
                        dplyr::select(all_of(to_select)) %>% 
                        unique() %>% 
                        left_join(program_scores_IPF %>%
                                        dplyr::select(program_name,
                                                      final_score),
                                  by = join_by("program_short" == "program_name")) %>% 
                        mutate(analysis = "DEA")
programs_both    <- gigatable %>% 
                        filter(!if_all(DEA_cols, is.na),                                                                    # Significant in at least one DEA method
                               FDR_lmm_phase < FDR_padj_thresh & abs(log2FC_lmm_phase) > log2FC_lmm_phase_thresh) %>%       # Significant in LMM
                        dplyr::select(all_of(to_select)) %>% 
                        unique() %>% 
                        left_join(program_scores_IPF %>%
                                        dplyr::select(program_name,
                                                      final_score),
                                  by = join_by("program_short" == "program_name")) %>% 
                        mutate(analysis = "Both")

plot_data <- bind_rows(list(programs_spectra, programs_dea, programs_both))

p <- plot_data %>% ggplot(aes(x = analysis, y = final_score)) +
                        geom_boxplot(outliers = FALSE) +      # Not showing outliers since datapoints are shown
                        geom_jitter() +
                        ggtitle("Biological relevance in IPF (final_score from cytopus relevance report) in programs found deregulated only by Spectra, only by DEA methods (any) or both.") +
                        ylab("Relevance score") +
                        theme_minimal()

ggsave(plot       = p,
       filename   = "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/methods_comparison/biological_relevace/SPECTRA_vs_DEA_vs_both.pdf",
       device     = "pdf",
       create.dir = TRUE,
       width      = 8,
       height     = 5)

```


