Filippo Gastaldello - 26/01/26

Analysis of the output of SPECTRA

```{r}
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggpubr)
library(ggrepel)
library(parallel)
library(Seurat)
library(eulerr)
library(patchwork)
library(ggplotify)
library(jsonlite)
library(tidyverse)
library(ggrastr)
library(msigdbr)
library(fgsea)
library(clusterProfiler)  
library(org.Hs.eg.db)
library(DOSE)
library(ReactomePA)
library(enrichplot)
```

# Data loading and preprocess
```{r}
# Define cell type annotation and gene sets to be used
annotation <- "manual"
annotation_column <- "manual_annotation_1"

# Cell scores quantify the activation of each gene program in every cell
cell_scores <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA/full/manual_annotation/Natri_et_al_full_manual_anno_cell_scores_93908.hpc3-head1.unitn.it.csv") %>% dplyr::rename("cell_id"="...1")
# Expression metadata are needed to match cell id to their cell type
metadata <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/Natri_et_al_metadata.csv")
# Only keep metadata for cells in the SPECTRA output
metadata <- metadata %>% filter(cell_id %in% cell_scores$cell_id)
# Add condition column to cell scores
cell_scores <- cell_scores %>% 
                        left_join(metadata %>% dplyr::select(cell_id, Diagnosis, Sample_Name, Phase, annotation_column)) %>%
                        relocate(Diagnosis, .before = cell_id) %>%
                        relocate(Phase, .after = Diagnosis) %>%
                        relocate(Sample_Name, .before = Diagnosis) %>% 
                        relocate(annotation_column, .after=Diagnosis)
colnames(cell_scores)[3] <- "annotation"
# Remove programs without a name
unknown_programs <- grep("\\d+$", colnames(cell_scores)[6:length(colnames(cell_scores))], value = TRUE)
cell_scores <- cell_scores %>% dplyr::select(-all_of(unknown_programs))
```
# Upnaired Welch's test (DAP)

Test IGP activation differences in case vs control, for each cell type

```{r}
res_welch <- lapply(unique(cell_scores$annotation),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% filter(annotation == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "t-statistic"=numeric(),
                                            "df"=numeric(), "p"=numeric(), "control_avg"=numeric(), "case_avg"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Try to compute the test, it might fail if not enough cells are active
                                test_res <- try(t.test(x = (celltype_cell_score %>% dplyr::filter(Diagnosis == "Control" & .[[program]]>0.001) %>% dplyr::select(all_of(program))),
                                                       y = (celltype_cell_score %>% dplyr::filter(Diagnosis == "IPF" & .[[program]]>0.001) %>% dplyr::select(all_of(program))),
                                                       var.equal = FALSE),
                                                       silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "t-statistic"=NA, "df"=NA,
                                                                  "p"=NA, "control_avg"=NA, "case_avg"=NA)
                                                       )
                                }else{
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "t-statistic"=test_res$statistic, "df"=test_res$parameter,
                                                                  "p"=test_res$p.value, "control_avg"=test_res$estimate[1], "case_avg"=test_res$estimate[2])
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_welch <- bind_rows(res_welch)

# Compute FDR to account for multiple testing
IGP_activation_welch <- IGP_activation_welch %>%
                                drop_na() %>% 
                                group_by(cell_type) %>% 
                                mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                                ungroup() %>% 
        relocate(FDR, .after = p)

# Compute two measure of effect size
#
# logFC <- log(case_avg/control_avg)
# delta_active <- active_cells_case_perc - active_cells_control_perc
IGP_activation_welch <- IGP_activation_welch %>% mutate(delta_active = active_cells_case_perc - active_cells_control_perc,
                                                        log2FC = log2(case_avg/control_avg))

if(! dir.exists(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/"))){dir.create(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/"), recursive = TRUE)}
write_csv(IGP_activation_welch, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/IPG_activation_welch_test.csv"))
```

## Boxplots - facet

Plot activation values in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
# Extract significant tests
welch_sig <- IGP_activation_welch %>% dplyr::filter(FDR <0.05 & (abs(log2FC) > 1))
# Flatten cell_scores for plotting
plot_data <- cell_scores %>% pivot_longer(6:length(colnames(cell_scores)), names_to = "program")
# Only keep data points representing active cells ( activation > 0.001) belonging to significant comparisons
plot_data <- plot_data %>% inner_join(welch_sig %>% dplyr::select(cell_type, program), by = join_by("annotation"=="cell_type", "program"=="program")) %>% dplyr::rename("cell_type"="annotation") %>% dplyr::filter(value>0.001)

p <- plot_data %>% ggplot(aes(x=program, y=value, fill=Diagnosis))+
                geom_boxplot()+
                theme(
                        axis.text.x = element_text(hjust=1, angle = 45, size = 5)
                )+
                facet_wrap(~cell_type, scale="free")
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/boxplots_DAP/boxplot_welch_significant_facet_celltype.pdf"), plot = p, device = "pdf", width = 18, height = 15, create.dir = TRUE)
```
## Boxplots - per celltype

Plot activation values in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
for (celltype in unique(plot_data$cell_type)) {
        
        p <- plot_data %>% filter(cell_type==celltype) %>% ggplot(aes(x=program, y=value, fill=Diagnosis)) +
                                                                geom_boxplot() +
                                                                labs(title = paste0("IGP with significant differences in activation between case and control in ",celltype," cells."),
                                                                     y = "Activation level"
                                                                ) +
                                                                theme(axis.text.x = element_text(hjust=1, angle=45, size=8)) +
                                                                stat_compare_means(
                                                                        aes(group = Diagnosis),
                                                                        method = "t.test",
                                                                        method.args = list(var.equal=FALSE),
                                                                        vjust = -1
                                                                ) +
                                                                stat_summary(
                                                                    fun.data = function(y) {
                                                                      return(data.frame(
                                                                        y = max(y),
                                                                        label = paste0("n= ",length(y))
                                                                      ))
                                                                    },
                                                                    geom = "text",
                                                                    position = position_dodge(width = 0.75),
                                                                    vjust = -0.5,
                                                                    size = 3
                                                                  )
                                                                                        
        
        ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation,"_annotation/welch_test/boxplots_DAP/boxplot_welch_significant_", gsub("/","_",celltype),".pdf"), plot = p, device = "pdf", width = 12, height = 8, create.dir = TRUE)
}
```
## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_welch %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=log2FC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Welch's t-test",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & log2FC>FC_thresh) | (FDR<FDR_thresh & log2FC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/volcano_welch_DAP.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```
# Unpaired Welch's test (expression)

Test for differences in gene expression, per cell-type, per program

```{r}
# Load seurat object for expression data and cytopus gene sets
data <- read_rds("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/Natri_et_al_clean.rds")
Idents(data) <- annotation_column
cytopus_gene_sets <- read_json("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/cytopus_gene_sets/knowledgebase_gene_sets_manual_annotation.json")

celltypes <- metadata %>% pull(any_of(annotation_column)) %>% unique()

res_welch <- lapply(celltypes, function(cell_type){
  
        # Subset gene expression to keep only current cell type
        celltype_exp <- GetAssayData(subset(data, idents = cell_type), assay = "RNA", layer = "data")
        celltype_exp <- t(as.matrix(celltype_exp))
        # Identify Control and IPF cells once per cell type
        ctrl_cells <- metadata %>% dplyr::filter(Diagnosis == "Control", .[[annotation_column]] == cell_type) %>% pull(cell_id)
        case_cells <- metadata %>% dplyr::filter(Diagnosis == "IPF", .[[annotation_column]] == cell_type) %>% pull(cell_id)
        # Ensure we only ask for cells that exist in the matrix
        ctrl_cells <- intersect(ctrl_cells, rownames(celltype_exp))
        case_cells <- intersect(case_cells, rownames(celltype_exp))
        # Extract gene_sets of current celltype
        celltype_gene_sets <- c(cytopus_gene_sets[[cell_type]], cytopus_gene_sets[["global"]])
        
        program_names <- names(celltype_gene_sets)
        
        test_results_list <- lapply(program_names, function(prog_name) {
        
                # Retrieve genes involved in the program
                program_genes <- unlist(celltype_gene_sets[[prog_name]])
                valid_genes <- intersect(program_genes, colnames(celltype_exp))
                
                # Skip if no valid genes to prevent matrix subsetting errors
                if(length(valid_genes) == 0) {
                        return(data.frame("cell_type"=cell_type, "program"=prog_name, "t-statistic"=NA, "df"=NA, "p"=NA, "control_avg"=NA, "case_avg"=NA))
                }
                
                expr_ctrl_vals <- as.vector(celltype_exp[ctrl_cells, valid_genes, drop = FALSE])
                expr_case_vals <- as.vector(celltype_exp[case_cells, valid_genes, drop = FALSE])
                
                # Try to compute the test
                test_res <- try(t.test(x = expr_ctrl_vals,
                                       y = expr_case_vals),
                                silent = TRUE)
                
                if (inherits(test_res, "try-error")) {
                        return(data.frame("cell_type"=cell_type, "program"=prog_name, "t-statistic"=NA, "df"=NA, "p"=NA, "control_avg"=NA, "case_avg"=NA))
                } else {
                        return(data.frame("cell_type"=cell_type, "program"=prog_name, "t-statistic"=test_res$statistic, "df"=test_res$parameter, "p"=test_res$p.value, "control_avg"=test_res$estimate[1], "case_avg"=test_res$estimate[2]))
                }
        })
        tests <- bind_rows(test_results_list)
        return(tests)
})

IGP_expression_welch <- bind_rows(res_welch)

# Compute FDR to account for multiple testing
IGP_expression_welch <- IGP_expression_welch %>%
                                drop_na() %>% 
                                group_by(cell_type) %>% 
                                mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                                ungroup() %>% 
        relocate(FDR, .after = p)

# logFC <- log(case_avg/control_avg)
IGP_expression_welch <- IGP_expression_welch %>% mutate(log2FC = log2(case_avg/control_avg))

write_csv(IGP_expression_welch, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/IPG_expression_welch_test.csv"))
```

## Boxplots - facet and per celltype

Plot expression distribution for all genes belonging to each program for each cell type in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
# Extract significant tests
welch_sig <- IGP_expression_welch %>% filter(FDR <0.05 & (abs(log2FC) > 1))

# One plot per type as well as facet
plots <- lapply(unique(welch_sig$cell_type), function(celltype){
        
        # Subset gene expression to keep only current cell type
        celltype_exp <- GetAssayData(subset(data, idents = celltype), assay = "RNA", layer = "data")
        celltype_exp <- t(as.matrix(celltype_exp))
        
        # For the current celltype, extract gene expression for eachy significant program
        plot_data <- lapply(welch_sig %>% dplyr::filter(cell_type == celltype) %>% pull(program), function(program_name){
                
                # Get genes involved in program (Check if program is cell specific or general)
                if (str_split_i(program_name, "_", 1) == "all") {
                        program_genes <- unlist(cytopus_gene_sets[["global"]][[program_name]])
                }else{
                        program_genes <- unlist(cytopus_gene_sets[[celltype]][[program_name]])
                }
                valid_genes <- intersect(program_genes, colnames(celltype_exp))
                # Get expression for these genes, pivot and set cell_id as column
                program_genes_exp <- celltype_exp[,valid_genes] %>% as.data.frame() %>% rownames_to_column(var = "cell_id")
                program_genes_exp <- program_genes_exp %>% pivot_longer(cols = colnames(program_genes_exp)[-1])
                # Add information about condition
                program_genes_exp <- program_genes_exp %>% left_join(metadata %>% dplyr::select(cell_id, Diagnosis))
                # Add program column
                program_genes_exp <- program_genes_exp %>% mutate(program = program_name)
        })
        plot_data <- bind_rows(plot_data)
        
        # Plot boxplot for this celltype
        p <- plot_data %>% ggplot(aes(x=program, y=log(value), fill=Diagnosis)) +
                                geom_boxplot() +
                                labs(title = paste0("IGP with significant differences in gene expression between case and control in ",celltype," cells."),
                                     y = "log(Gene expression)"
                                ) +
                                theme(axis.text.x = element_text(hjust=1, angle=45, size=8)) +
                                stat_compare_means(
                                        aes(group = Diagnosis),
                                        method = "t.test",
                                        label = "p.signif",
                                        method.args = list(var.equal=FALSE),
                                        vjust = -1,
                                        size = 3
                                ) +
                                stat_summary(
                                    fun.data = function(y) {
                                      return(data.frame(
                                        y = max(y),
                                        label = paste0("n= ",length(y))
                                      ))
                                    },
                                    geom = "text",
                                    position = position_dodge(width = 0.75),
                                    vjust = -0.5,
                                    size = 3
                                  ) +
                                theme_minimal()
                                                                                        
        
        ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation,"_annotation/welch_test/boxplots_DEG/boxplot_welch_significant_", gsub("/","_",celltype),".pdf"), plot = p, device = "pdf", width = 10, height = 7, create.dir = TRUE)
        
        return(p)
})

combined_plot <- wrap_plots(plots)
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation ,"_annotation/welch_test/boxplots_DEG/boxplot_welch_significant_facet.pdf"), plot = combined_plot, width = 25, height = 25, create.dir = TRUE)
```

## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_expression_welch %>% 
                mutate(tag = paste0(cell_type, "-", program),
                       diff_expressed = ifelse(FDR<FDR_thresh & log2FC>FC_thresh, "UP", "NO"),
                       diff_expressed = ifelse(FDR<FDR_thresh & log2FC< -FC_thresh, "DOWN", diff_expressed))

p <- plot_data %>% ggplot(aes(x=log2FC, y=-log10(FDR), color = diff_expressed)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Repressed", "Not significant", "Overexpressed")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Welch's t-test",
                        subtitle = "comparison between gene expression in case vs control, for each program in each cell type",
                        x = "log2(expression FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & log2FC>FC_thresh) | (FDR<FDR_thresh & log2FC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/welch_test/volcano_welch_DEG.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```

# Wilcoxon-Mann-Whitney test (DAP)

Test IGP activation differences in case vs control, for each cell type

```{r}
library(dplyr)
library(tidyr)

res_WMW <- lapply(unique(cell_scores$annotation), function(cell_type){
  
        # Subset cell scores to keep only current cell type
        celltype_cell_score <- cell_scores %>% dplyr::filter(annotation == cell_type)
        
        # Store total cells for percentages
        n_total_control <- sum(celltype_cell_score$Diagnosis == "Control")
        n_total_ipf     <- sum(celltype_cell_score$Diagnosis == "IPF")
        
        # Use a list to store loop results
        tests_list <- list()
        
        programs <- colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]
        
        for (program in programs) {
        
                # Extract active cells
                vec_control <- celltype_cell_score %>% dplyr::filter(Diagnosis == "Control" & .[[program]] > 0.001) %>% pull(all_of(program))
                vec_ipf     <- celltype_cell_score %>% dplyr::filter(Diagnosis == "IPF" & .[[program]] > 0.001) %>% pull(all_of(program))
                
                n_active_control <- length(vec_control)
                n_active_case    <- length(vec_ipf)
                
                perc_active_control <- round((n_active_control / n_total_control) * 100, 2)
                perc_active_case    <- round((n_active_case / n_total_ipf) * 100, 2)
                
                # Try to compute the test
                test_res <- try(wilcox.test(x = vec_control, y = vec_ipf, paired = FALSE), silent = TRUE)
                
                if (inherits(test_res, "try-error")) {
                        stat <- NA
                        pval <- NA
                }else{
                        stat <- test_res$statistic
                        pval <- test_res$p.value
                }
                
                # Store results, including the active sample sizes
                tests_list[[program]] <- data.frame(
                        cell_type = cell_type, 
                        program = program, 
                        active_cells_control_perc = perc_active_control,
                        active_cells_case_perc = perc_active_case, 
                        n_active_control = n_active_control, 
                        n_active_case = n_active_case,       
                        statistic = stat, 
                        p = pval
                )
        }
        return(bind_rows(tests_list))
})

# Bind all cell types together
IGP_activation_WMW <- bind_rows(res_WMW)

# Compute FDR and Effect Size
IGP_activation_WMW <- IGP_activation_WMW %>%
  drop_na(p) %>% 
  group_by(cell_type) %>% 
  mutate(FDR = p.adjust(p, method = "fdr")) %>% 
  ungroup() %>% 
  relocate(FDR, .after = p) %>%
  # Apply the Rank-Biserial Formula using the sample sizes
  mutate(effect_size = abs(1 - (2 * statistic) / (n_active_control * n_active_case)))

if(! dir.exists(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/"))){dir.create(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/"), recursive = TRUE)}
write_csv(IGP_activation_WMW, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/IPG_activation_WMW_test.csv"))
```

## Boxplots - facet

Plot activation values in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
# Extract significant tests
WMW_sig <- IGP_activation_WMW %>% dplyr::filter(FDR <0.05 & (effect_size > 0.3))
# Flatten cell_scores for plotting
plot_data <- cell_scores %>% pivot_longer(6:length(colnames(cell_scores)), names_to = "program")
# Only keep data points representing active cells ( activation > 0.001) belonging to significant comparisons
plot_data <- plot_data %>% inner_join(WMW_sig %>% dplyr::select(cell_type, program), by = join_by("annotation"=="cell_type", "program"=="program")) %>% dplyr::rename("cell_type"="annotation") %>% dplyr::filter(value>0.001)

p <- plot_data %>% ggplot(aes(x=program, y=value, fill=Diagnosis))+
                geom_boxplot()+
                theme(
                        axis.text.x = element_text(hjust=1, angle = 45, size = 5)
                )+
                facet_wrap(~cell_type, scale="free")
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/boxplots_DAP/boxplot_wilcoxon_significant_facet_celltype.pdf"), plot = p, device = "pdf", width = 18, height = 15, create.dir = TRUE)
```
## Boxplots - per celltype

Plot activation values in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
for (celltype in unique(plot_data$cell_type)) {
        
        p <- plot_data %>% dplyr::filter(cell_type==celltype) %>% ggplot(aes(x=program, y=value, fill=Diagnosis)) +
                                                                geom_boxplot() +
                                                                labs(title = paste0("IGP with significant differences in activation between case and control in ",celltype," cells."),
                                                                     y = "Activation level"
                                                                ) +
                                                                theme(axis.text.x = element_text(hjust=1, angle=45, size=8)) +
                                                                stat_compare_means(
                                                                        aes(group = Diagnosis),
                                                                        method = "t.test",
                                                                        method.args = list(var.equal=FALSE),
                                                                        vjust = -1
                                                                ) +
                                                                stat_summary(
                                                                    fun.data = function(y) {
                                                                      return(data.frame(
                                                                        y = max(y),
                                                                        label = paste0("n= ",length(y))
                                                                      ))
                                                                    },
                                                                    geom = "text",
                                                                    position = position_dodge(width = 0.75),
                                                                    vjust = -0.5,
                                                                    size = 3
                                                                  )
                                                                                        
        
        ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation,"_annotation/wilcoxon_test/boxplots_DAP/boxplot_wilcoxon_significant_", gsub("/", "_", celltype),".pdf"), plot = p, device = "pdf", width = 12, height = 8, create.dir = TRUE)
}
```
## Scatter plot FDR vs effect size
```{r}
# create tag for labels and deregulation column
effect_size_thresh_medium <- 0.3
effect_size_thresh_high <- 0.5
FDR_thresh <- 0.05

plot_data <- IGP_activation_WMW %>% 
                mutate(tag            = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       effect         = ifelse(FDR<FDR_thresh & effect_size > effect_size_thresh_medium, "Medium effect size", "Small effect size"),
                       effect         = ifelse(FDR<FDR_thresh & effect_size > effect_size_thresh_high, "High effect size", effect),
                       effect         = as.factor(effect))

p <- plot_data %>% ggplot(aes(x=effect_size, y=-log10(FDR), color = effect)) +
                        geom_point(size = 1) +
                        scale_color_manual(values = c("High effect size" = "red", "Medium effect size" = "purple", "Small effect size" = "grey")) +
                        geom_vline(xintercept = 0.3, col = 'grey', linetype = 'dashed') +
                        geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                        labs(
                                title    = "Wilcoxon rank sum test",
                                subtitle = "comparison between activation in case vs control, for each program in each cell type",
                                x        = "Effect size") +
                        geom_label_repel(
                                data = plot_data %>% filter((FDR<FDR_thresh & effect_size>effect_size_thresh_high)),
                                aes(label = tag),
                                size = 2.5,           
                                max.overlaps = Inf,   
                                box.padding = 0.5,
                                point.padding = 0.2,
                                min.segment.length = 0,
                                nudge_x = .15,
                                nudge_y = .5,) +
                        theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/scatterplot_wilcoxon_DAP.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```
# Wilcoxon-Mann-Whitney test (expression)

Test for differences in gene expression, per cell-type, per program

```{r}
celltypes <- metadata %>% pull(any_of(annotation_column)) %>% unique()

res_welch <- lapply(celltypes, function(cell_type){
  
        # Subset gene expression to keep only current cell type
        celltype_exp <- GetAssayData(subset(data, idents = cell_type), assay = "RNA", layer = "data")
        celltype_exp <- t(as.matrix(celltype_exp))
        
        # Identify Control and IPF cells once per cell type
        ctrl_cells <- metadata %>% dplyr::filter(Diagnosis == "Control", .[[annotation_column]] == cell_type) %>% pull(cell_id)
        case_cells <- metadata %>% dplyr::filter(Diagnosis == "IPF", .[[annotation_column]] == cell_type) %>% pull(cell_id)
        
        # Ensure we only ask for cells that exist in the matrix
        ctrl_cells <- intersect(ctrl_cells, rownames(celltype_exp))
        case_cells <- intersect(case_cells, rownames(celltype_exp))
        
        # Extract gene_sets of current celltype
        celltype_gene_sets <- c(cytopus_gene_sets[[cell_type]], cytopus_gene_sets[["global"]])
        program_names <- names(celltype_gene_sets)
        
        test_results_list <- lapply(program_names, function(prog_name) {
                # Retrieve genes involved in the program
                program_genes <- unlist(celltype_gene_sets[[prog_name]])
                valid_genes   <- intersect(program_genes, colnames(celltype_exp))
                
                # Skip if no valid genes to prevent matrix subsetting errors
                if(length(valid_genes) == 0) {
                        return(data.frame("cell_type"=cell_type,
                                          "program"=prog_name,
                                          "statistic"=NA,
                                          "p"=NA,
                                          "ctrl_cells_count" = NA,
                                          "case_cells_count"=NA))
                }
                
                expr_ctrl_vals <- as.vector(celltype_exp[ctrl_cells, valid_genes, drop = FALSE])
                expr_case_vals <- as.vector(celltype_exp[case_cells, valid_genes, drop = FALSE])
                
                # Try to compute the test
                test_res <- try(wilcox.test(x = expr_ctrl_vals, y = expr_case_vals, paired = FALSE))
                
                if (inherits(test_res, "try-error")) {
                        return(data.frame("cell_type"=cell_type,
                                          "program"=prog_name,
                                          "statistic"=NA,
                                          "p"=NA,
                                          "ctrl_cells_count"=NA,
                                          "case_cells_count"=NA))
                } else {
                        return(data.frame("cell_type"=cell_type,
                                          "program"=prog_name,
                                          "statistic"=test_res$statistic,
                                          "p"=test_res$p.value,
                                          "ctrl_cells_count"=length(expr_ctrl_vals),
                                          "case_cells_count"=length(expr_case_vals)))
                }
        })
        tests <- bind_rows(test_results_list)
        return(tests)
})

IGP_expression_WMW <- bind_rows(res_welch)

# Compute FDR to account for multiple testing
IGP_expression_WMW <- IGP_expression_WMW %>%
                                drop_na() %>% 
                                group_by(cell_type) %>% 
                                mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                                ungroup() %>% 
                                relocate(FDR, .after = p) %>% 
                                # Apply the Rank-Biserial Formula using the sample sizes
                                mutate(effect_size = abs(1 - (2 * statistic) / (as.numeric(ctrl_cells_count) * as.numeric(case_cells_count))))

write_csv(IGP_expression_WMW, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/IPG_expression_WMW_test.csv"))
```

## Boxplots - facet and per celltype

Plot expression distribution for all genes belonging to each program for each cell type in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
# Extract significant tests
WMW_sig <- IGP_expression_WMW %>% dplyr::filter(FDR <0.05 & effect_size > 0.3)

# One plot per type as well as facet
plots <- lapply(unique(WMW_sig$cell_type), function(celltype){
        
        # Subset gene expression to keep only current cell type
        celltype_exp <- GetAssayData(subset(data, idents = celltype), assay = "RNA", layer = "data")
        celltype_exp <- t(as.matrix(celltype_exp))
        
        # For the current celltype, extract gene expression for eachy significant program
        plot_data <- lapply(WMW_sig %>% dplyr::filter(cell_type == celltype) %>% pull(program), function(program_name){
                
                # Get genes involved in program (Check if program is cell specific or general)
                if (str_split_i(program_name, "_", 1) == "all") {
                        program_genes <- unlist(cytopus_gene_sets[["global"]][[program_name]])
                }else{
                        program_genes <- unlist(cytopus_gene_sets[[celltype]][[program_name]])
                }
                valid_genes <- intersect(program_genes, colnames(celltype_exp))
                # Get expression for these genes, pivot and set cell_id as column
                program_genes_exp <- celltype_exp[,valid_genes] %>% as.data.frame() %>% rownames_to_column(var = "cell_id")
                program_genes_exp <- program_genes_exp %>% pivot_longer(cols = colnames(program_genes_exp)[-1])
                # Add information about condition
                program_genes_exp <- program_genes_exp %>% left_join(metadata %>% dplyr::select(cell_id, Diagnosis))
                # Add program column
                program_genes_exp <- program_genes_exp %>% mutate(program = program_name)
        })
        plot_data <- bind_rows(plot_data)
        
        # Plot boxplot for this celltype
        p <- plot_data %>% ggplot(aes(x=program, y=log(value), fill=Diagnosis)) +
                                geom_boxplot() +
                                labs(title = paste0("IGP with significant differences in gene expression between case and control in ",celltype," cells."),
                                     y = "log(Gene expression)"
                                ) +
                                theme(axis.text.x = element_text(hjust=1, angle=45, size=8)) +
                                stat_compare_means(
                                        aes(group = Diagnosis),
                                        method = "t.test",
                                        label = "p.signif",
                                        method.args = list(var.equal=FALSE),
                                        vjust = -1,
                                        size = 3
                                ) +
                                stat_summary(
                                    fun.data = function(y) {
                                      return(data.frame(
                                        y = max(y),
                                        label = paste0("n= ",length(y))
                                      ))
                                    },
                                    geom = "text",
                                    position = position_dodge(width = 0.75),
                                    vjust = -0.5,
                                    size = 3
                                  ) +
                                theme_minimal()
                                                                                        
        
        ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation,"_annotation/wilcoxon_test/boxplots_DEG/boxplot_wilcoxon_significant_", gsub("/", "_", celltype),".pdf"), plot = p, device = "pdf", width = 10, height = 7, create.dir = TRUE)
        
        return(p)
})

combined_plot <- wrap_plots(plots)
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation ,"_annotation/wilcoxon_test/boxplots_DEG/boxplot_wilcoxon_significant_facet.pdf"), plot = combined_plot, width = 25, height = 25, create.dir = TRUE)
```

## Scatter plot FDR vs effect size
```{r}
# create tag for labels and deregulation column
effect_size_thresh_medium <- 0.3
effect_size_thresh_high <- 0.5
FDR_thresh <- 0.05

plot_data <- IGP_expression_WMW %>% 
                mutate(tag            = paste0(cell_type, "-", program),
                       effect         = ifelse(FDR<FDR_thresh & effect_size > effect_size_thresh_medium, "Medium effect size", "Small effect size"),
                       effect         = ifelse(FDR<FDR_thresh & effect_size > effect_size_thresh_high, "High effect size", effect),
                       effect         = as.factor(effect))

p <- plot_data %>% ggplot(aes(x=effect_size, y=-log10(FDR), color = effect)) +
                        geom_point(size = 1) +
                        scale_color_manual(values = c("High effect size" = "red", "Medium effect size" = "purple", "Small effect size" = "grey")) +
                        geom_vline(xintercept = 0.3, col = 'grey', linetype = 'dashed') +
                        geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                        labs(
                                title    = "Wilcoxon rank sum test",
                                subtitle = "comparison between gene expression in case vs control, for each program in each cell type",
                                x        = "Effect size") +
                        geom_label_repel(
                                data = plot_data %>% filter((FDR<FDR_thresh & effect_size>effect_size_thresh_medium)),
                                aes(label = tag),
                                size = 2.5,           
                                max.overlaps = Inf,   
                                box.padding = 0.5,
                                point.padding = 0.2,
                                min.segment.length = 0,
                                nudge_x = .15,
                                nudge_y = .5,) +
                        theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/wilcoxon_test/scatterplot_wilcoxon_DEG.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```

# Linear Mixed Model (DAP)

In Welch's test each cell is seen as an inpdendent measurement, ingoring the possible collinearity of cells belonging to the same patient. Linear Mixed Models correct for this modeling the patient id as a random effect.

```{r message=FALSE, warning=FALSE}
res_lmm <- lapply(unique(cell_scores$annotation),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% dplyr::filter(annotation == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "term"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "avg_control_activation"=numeric(), "avg_case_activation"= numeric(), "estimate"=numeric(),
                                            "std.err"=numeric(), "statistic"=numeric(), "p"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        dplyr::filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        dplyr::filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Compute average activation in case vs control
                                avg_control_activation <- celltype_cell_score %>%
                                                                dplyr::filter(Diagnosis == "Control") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                avg_case_activation <- celltype_cell_score %>%
                                                                dplyr::filter(Diagnosis == "IPF") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                # Try to compute the test, it might fail if not enough cells are active
                                data <- celltype_cell_score %>% dplyr::select(Diagnosis, Sample_Name, all_of(program)) %>% dplyr::filter(.[[program]]>0.001)
                                colnames(data) <- c("Diagnosis", "Sample_Name", "Program")
                                test_res <- try(lmer(Program ~ Diagnosis + (1|Sample_Name), data = data),
                                                silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, term = NA, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "avg_control_activation"=avg_control_activation,
                                                                  "avg_case_activation"=avg_case_activation, "estimate"=NA, "std.err"=NA,"statistic"=NA, "p"=NA)
                                                       )
                                }else{
                                        test_res <- tidy(test_res, effects = "fixed")[2,]
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "term"=test_res$term,
                                                                  "active_cells_control_perc"=perc_active_cells_control, "active_cells_case_perc"=perc_active_cells_case,
                                                                  "avg_control_activation"=avg_control_activation,"avg_case_activation"=avg_case_activation,
                                                                  "estimate"=test_res$estimate, "std.err"=test_res$std.error, "statistic"=test_res$statistic, "p"=test_res$p.value)
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_lmm <- bind_rows(res_lmm)
IGP_activation_lmm <- IGP_activation_lmm %>% 
                        drop_na() %>% 
                        group_by(cell_type) %>% 
                        mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                        ungroup()
# Compute effect size as activationFC log2((avg_control_activation + estimate)/avg_case_activation)
IGP_activation_lmm <- IGP_activation_lmm %>% mutate(log2FC = log2((avg_control_activation+estimate)/avg_control_activation))

if(! dir.exists(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/lmm/"))){dir.create(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/lmm/"), recursive = TRUE)}
write_csv(IGP_activation_lmm, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/lmm/IPG_activation_lmm.csv"))
```

## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_lmm %>%
                dplyr::filter(!log2FC=="NaN" ) %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=log2FC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Linear Mixed Model - Sample ID as random effect",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & log2FC>FC_thresh) | (FDR<FDR_thresh & log2FC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/lmm/volcano_lmm.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```


# Linear Mixed Model - cell cycle phase correction (DAP)

```{r message=FALSE, warning=FALSE}
res_lmm <- lapply(unique(cell_scores$annotation),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% dplyr::filter(annotation == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "term"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "avg_control_activation"=numeric(), "avg_case_activation"= numeric(), "estimate"=numeric(),
                                            "std.err"=numeric(), "statistic"=numeric(), "p"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        dplyr::filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        dplyr::filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Compute average activation in case vs control
                                avg_control_activation <- celltype_cell_score %>%
                                                                dplyr::filter(Diagnosis == "Control") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                avg_case_activation <- celltype_cell_score %>%
                                                                dplyr::filter(Diagnosis == "IPF") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                # Try to compute the test, it might fail if not enough cells are active
                                data <- celltype_cell_score %>% dplyr::select(Diagnosis, Sample_Name, Phase, all_of(program)) %>% dplyr::filter(.[[program]]>0.001)
                                colnames(data) <- c("Diagnosis", "Sample_Name", "Phase", "Program")
                                test_res <- try(lmer(Program ~ Diagnosis + Phase + (1|Sample_Name), data = data),
                                                silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, term = NA, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "avg_control_activation"=avg_control_activation,
                                                                  "avg_case_activation"=avg_case_activation, "estimate"=NA, "std.err"=NA, "statistic"=NA, "p"=NA)
                                                       )
                                }else{
                                        test_res <- tidy(test_res, effects = "fixed")[2:3,]
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "term"=test_res$term,
                                                                  "active_cells_control_perc"=perc_active_cells_control, "active_cells_case_perc"=perc_active_cells_case,
                                                                  "avg_control_activation"=avg_control_activation, "avg_case_activation"=avg_case_activation,
                                                                  "estimate"=test_res$estimate, "std.err"=test_res$std.error, "statistic"=test_res$statistic, "p"=test_res$p.value)
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_lmm_phase_corrected <- bind_rows(res_lmm)
IGP_activation_lmm_phase_corrected <- IGP_activation_lmm_phase_corrected %>% 
                        drop_na() %>% 
                        group_by(cell_type) %>% 
                        mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                        ungroup()
# Compute effect size as activationFC log2((avg_control_activation + estimate)/avg_case_activation)
IGP_activation_lmm_phase_corrected <- IGP_activation_lmm_phase_corrected %>% mutate(log2FC = log2((avg_control_activation+estimate)/avg_control_activation))

write_csv(IGP_activation_lmm_phase_corrected, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/lmm/IPG_activation_lmm_phase_corrected.csv"))
```

## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_lmm_phase_corrected %>%
                dplyr::filter(!log2FC=="NaN",
                       !str_detect(term, "Phase")) %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=log2FC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Linear Mixed Model - Sample ID as random effect, correcting for cell cycle phase",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & log2FC>FC_thresh) | (FDR<FDR_thresh & log2FC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation,"_annotation/lmm/volcano_lmm_phase_corrected.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```

# Comparison DEG vs DAP

Compare gene deregulation (DEG) with differentialy activated program (DAP)

## Perform DE analysis

- Load original Seurat object
- perform differential analysis per cell type across condition


### SC
```{r}
# Create new metadata column to use as comparison in DE analysis
data@meta.data <- data@meta.data %>% 
                        mutate(celltype_condition = paste0(.[[annotation_column]], "-", Diagnosis))
Idents(data) <- "celltype_condition"
```

Run DE in parallel for each cell type
```{r}
cores <- 10

res_DE <- mclapply(unique(data@meta.data[,annotation_column]),
                function(cell_type){
                        return(FindMarkers(data, ident.1 = paste0(cell_type, "-IPF"), ident.2 = paste0(cell_type, "-Control")) %>%
                                       mutate(celltype = cell_type,
                                              FDR = p.adjust(p_val, method = "BH")) %>%
                                       rownames_to_column(var = "gene"))
                },
                mc.cores = cores,
                mc.cleanup = TRUE)
# Combine results from different cell types
DEGs <- bind_rows(res_DE)
```

#### Define Log2FC threshold
```{r}
FDR_thresh <- 0.01

# Check FDR quantiles per cell type to set thresholds (only considering test below FDR threshold)
res_quant <- lapply(unique(data@meta.data[,annotation_column]),
                        function(cell_type){
                                FC_dist <- ecdf(DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh) %>% pull(avg_log2FC) %>% abs())
                                quant <- data.frame(celltype = cell_type,
                                                    genes_below_FDR = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh) %>% nrow(),
                                                    FC_0.2_perc  = round((1 - FC_dist(0.2))*100, 2),
                                                    FC_0.2_count = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh & abs(avg_log2FC) > 0.2) %>% nrow(),
                                                    FC_0.5_perc  = round((1 - FC_dist(0.5))*100, 2),
                                                    FC_0.5_count = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh & abs(avg_log2FC) > 0.5) %>% nrow(),
                                                    FC_0.7_perc  = round((1 - FC_dist(0.7))*100, 2),
                                                    FC_0.7_count = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh & abs(avg_log2FC) > 0.7) %>% nrow(),
                                                    FC_1_perc    = round((1 - FC_dist(1))*100, 2),
                                                    FC_1_count   = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh & abs(avg_log2FC) > 1) %>% nrow(),
                                                    FC_1.5_perc  = round((1 - FC_dist(1.5))*100, 2),
                                                    FC_1.5_count   = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh & abs(avg_log2FC) > 1.5) %>% nrow(),
                                                    FC_2_perc  = round((1 - FC_dist(2))*100, 2),
                                                    FC_2_count   = DEGs %>% filter(cell_type == celltype & FDR < FDR_thresh & abs(avg_log2FC) > 2) %>% nrow())
                                                
                        })
# Combine results from different cell types
FDR_quantiles <- bind_rows(res_quant)
rownames(FDR_quantiles) <- seq(1,nrow(FDR_quantiles))

# Choose Log2FC threshold
FC_thresh  <- 1.5
# Retrieve the second lowest p value (the lowest might be zero in sc DE analysis) to use as pseudo-count for rank metric
lowest_p <- (DEGs$FDR %>% unique() %>% sort())[2]
# Label deregulated genes and add rank metric (-log10(FDR)*log2FC)
DEGs <- DEGs %>% mutate(is_sig = ifelse(FDR < FDR_thresh & abs(avg_log2FC) > FC_thresh, TRUE, FALSE),
                        class  = ifelse(avg_log2FC > 0, "UP", "DOWN"),
                        rank   = -log10(FDR+lowest_p)*avg_log2FC) # pseudo count to avoid computing log(0) in case of inflated pvalues
sig_DEGs_stats <- FDR_quantiles %>% dplyr::select(celltype, genes_below_FDR, paste0("FC_", FC_thresh, "_perc"),paste0("FC_", FC_thresh, "_count"))
```

### PSEUDO-BULK

Pseudo bulk based on sample - condition - cell type
```{r}
data_pb <- AggregateExpression(data, assays = "RNA", return.seurat = TRUE, group.by = c("Sample_Name", "Diagnosis", annotation_column))
# Create new idents
data_pb$celltype_condition <- paste0(data_pb@meta.data[, annotation_column], "-", data_pb$Diagnosis)
Idents(data_pb) <- "celltype_condition"


```

Run DE for each cell type
```{r message=FALSE}
res_DE_pb <- lapply(unique(data_pb@meta.data[,annotation_column]),
                    function(cell_type){
                            return(FindMarkers(data_pb, ident.1 = paste0(cell_type, "-IPF"), ident.2 = paste0(cell_type, "-Control"), test.use = "DESeq2") %>%
                                           mutate(celltype = cell_type,
                                                  FDR = p.adjust(p_val, method = "BH")) %>%
                                           rownames_to_column(var = "gene")
                            )
                            }
                    )
# Combine results from different cell types
DEGs_pb <- bind_rows(res_DE_pb)
# Label deregulated genes and add rank metric (-log10(FDR)*log2FC)
DEGs_pb <- DEGs_pb %>% mutate(is_sig = ifelse(FDR < FDR_thresh & abs(avg_log2FC) > FC_thresh, TRUE, FALSE),
                              is_sig = ifelse(is.na(p_val), FALSE, is_sig),
                              class  = ifelse(avg_log2FC > 0, "UP", "DOWN"),
                              rank   = -log10(FDR)*avg_log2FC)
```

### Merge DEGs from sc and pseudo-bulk and save all 3 tables
```{r}
total_DEGs <- DEGs %>% full_join(DEGs_pb, by = c("gene", "celltype"), suffix = c("_sc", "_pb"))
# Create flag for singificance both in sc and pb
total_DEGs <- total_DEGs %>% mutate(both_sig = ifelse(is_sig_sc == TRUE & is_sig_pb == TRUE, TRUE, FALSE))


if(! dir.exists(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/"))){dir.create(paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/"), recursive = TRUE)}
# Save all 3
write_csv(DEGs, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/sc_DEGs.csv"))
write_csv(DEGs_pb, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/pb_DEGs.csv"))
write_csv(total_DEGs, file = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/total_DEGs.csv"))
```

## Check overlap of DEGs in sc vs pseudo bulk
```{r}
plot_list <- lapply(unique(data@meta.data[[annotation_column]]), function(cell_type) {
  
                  sc_genes <- DEGs %>% dplyr::filter(is_sig == TRUE, celltype == cell_type) %>% pull(gene)
                  pb_genes <- DEGs_pb %>% dplyr::filter(is_sig == TRUE, celltype == cell_type) %>% pull(gene)
                  
                  fit <- euler(list("sc" = sc_genes, "Pseudo-bulk" = pb_genes))
                  
                  # Plot ven
                  p1 <- as.ggplot(plot(fit, quantities = TRUE, fills = c("skyblue", "pink"), alpha = 0.5))
                  # Combine with correlation of FC in sc vs pb
                  p2 <- total_DEGs %>%
                        dplyr::filter(both_sig, celltype == cell_type) %>% 
                        dplyr::select(avg_log2FC_sc, avg_log2FC_pb) %>%
                        drop_na() %>%
                        ggplot(aes(x = avg_log2FC_sc, y = avg_log2FC_pb)) +
                                geom_point_rast(size = 0.1) +
                                geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE, size = 0.25) +
                                labs(title = cell_type) +
                                theme_minimal()
                  p <- wrap_plots(c(p1, p2)) + plot_annotation(title = cell_type)
})

combined_plot <- wrap_plots(plot_list) + 
  plot_annotation(title    = "Overlap between DEGs found in scDE and pseudo-bulkDE analyses",
                  subtitle = "Scatterplot showing logFC correlation between genes significantly altered (FDR<0.05 and log2FC>1) in both analyses (venn intersection)" )

ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation ,"_annotation/DE/sc_vs_pseudobulk_overlap.pdf"), plot = combined_plot, width = 25, height = 11, create.dir = TRUE)
```


## Check correlation between log2FCs and FDRs

FDRs
```{r}
# Significant in one of the 2
p1 <- total_DEGs %>%
        dplyr::filter(is_sig_sc| is_sig_pb) %>% 
        dplyr::select(FDR_sc, FDR_pb) %>%
        drop_na() %>%
        dplyr::filter(!FDR_sc == 0) %>% 
        ggplot(aes(x = -log(FDR_sc), y = -log(FDR_pb))) +
                geom_point_rast(size = 0.01, alpha = 0.07, raster.dpi = 400) +
                geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE, size = 0.25) +
                labs(title = "Significant in at least one of the two analyses") +
                theme_minimal()

# Significant in both
p2 <- total_DEGs %>%
        dplyr::filter(both_sig) %>% 
        dplyr::select(FDR_sc, FDR_pb) %>%
        drop_na() %>%
        dplyr::filter(!FDR_sc == 0) %>% 
        ggplot(aes(x = -log(FDR_sc), y = -log(FDR_pb))) +
                geom_point_rast(size = 0.01, alpha = 0.07, raster.dpi = 400) +
                geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE, size = 0.25) +
                labs(title = "Significant in both analyses") +
                theme_minimal()
p <- wrap_plots(p1, p2) + plot_annotation(title = "Correlation between single cell and pseudobulk FDRs of differential analysis",
                                          subtitle = "(Removed genes where single cell FDR was zero)")
p

ggsave(p, filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/FDR_sc_pb_correlation.pdf"), device = "pdf", width = 10, height = 5, create.dir = TRUE)
```

log2FC
```{r}
# Significant in one of the 2
p1 <- total_DEGs %>%
        dplyr::filter(is_sig_sc | is_sig_pb) %>% 
        dplyr::select(avg_log2FC_sc, avg_log2FC_pb) %>%
        drop_na() %>% 
        ggplot(aes(x = avg_log2FC_sc, y = avg_log2FC_pb)) +
                geom_point_rast(size = 0.01, alpha = 0.07, raster.dpi = 400) +
                geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE, size = 0.25) +
                labs(title = "Significant in at least one of the two analyses") +
                theme_minimal()

# Significant in both
p2 <- total_DEGs %>%
        dplyr::filter(both_sig) %>% 
        dplyr::select(avg_log2FC_sc, avg_log2FC_pb) %>%
        drop_na() %>%
        ggplot(aes(x = avg_log2FC_sc, y = avg_log2FC_pb)) +
                geom_point_rast(size = 0.01, alpha = 0.07, raster.dpi = 400) +
                geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE, size = 0.25) +
                labs(title = "Significant in both analyses") +
                theme_minimal()
p <- wrap_plots(p1, p2) + plot_annotation(title = "Correlation between single cell and pseudobulk Log2FC of differential analysis")
p

ggsave(p, filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/",annotation,"_annotation/DE/log2FC_sc_pb_correlation.pdf"), device = "pdf", width = 10, height = 5, create.dir = TRUE)
```

# DEGs enrichment

## GSEA

### Prepare ranks
 
Fetch GO and REACTOME and "Hallmark" gene sets from msigDB
```{r}
# Fetch GO Biological Process 
go_bp_df   <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP")
go_bp_list <- split(x = go_bp_df$gene_symbol, f = go_bp_df$gs_name)

# Fetch Reactome 
reactome_df   <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME")
reactome_list <- split(x = reactome_df$gene_symbol, f = reactome_df$gs_name)

# Create cytopus_list
cytopus_list        <- unlist(cytopus_gene_sets, recursive = FALSE)
names(cytopus_list) <- str_split_i(names(cytopus_list), "\\.",2) 
cytopus_list        <- cytopus_list[!duplicated(names(cytopus_list))]
cytopus_list        <- lapply(cytopus_list, function(x){unlist(x)})

gene_sets       <- list("GO"=go_bp_list, "REACTOME"=reactome_list, "CYTOPUS"=cytopus_list)
gene_sets_names <- c("GO", "REACTOME", "CYTOPUS")
```

### Run gsea
```{r}
# Output path
base_out_path <- paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation, "_annotation/enrichment/gsea/")

# Define configurations for each analysis
analysis_configs <- list(
  sc = list(
    sig_col    = "is_sig_sc",
    rank_col   = "rank_sc",
    title_desc = "single cell DE analysis",
    out_folder = "sc"
  ),
  pb = list(
    sig_col    = "is_sig_pb",
    rank_col   = "rank_pb",
    title_desc = "pseudo-bulk DE analysis",
    out_folder = "pb"
  )
)

# For each celltype run gsea for sc and pb DEGs 
for (cell_type in unique(total_DEGs$celltype)) {
        
        # Iterate over sc and pb analyses
        for (ana_name in names(analysis_configs)) {
                
                # Get configuration for the current analysis
                cfg <- analysis_configs[[ana_name]]
                
                # Create rank for this celltype based on the configuration
                celltype_DEGs <- total_DEGs %>% 
                dplyr::filter(celltype == cell_type, !is.na(.data[[cfg$sig_col]])) %>% 
                dplyr::arrange(desc(.data[[cfg$rank_col]]))
                
                rank        <- celltype_DEGs[[cfg$rank_col]]
                names(rank) <- celltype_DEGs$gene
                rank        <- rank[!is.na(rank)]
                for (gene_set in gene_sets_names) {
                        
                        set.seed(42)
                        fgsea_results <- fgsea(pathways = gene_sets[[gene_set]], 
                                               stats    = rank,
                                               minSize  = 5,
                                               maxSize  = 500)
                        
                        sig_results   <- fgsea_results[padj < 0.05]
                        top_up        <- head(sig_results[order(-NES)]$pathway, 10)
                        top_down      <- head(sig_results[order(NES)]$pathway, 10)
                        top_pathways  <- unique(c(top_up, top_down))
                        
                        if (!is_empty(top_pathways)) {
                                p <- as.ggplot(plotGseaTable(gene_sets[[gene_set]][top_pathways],
                                                             rank, 
                                                             fgsea_results,
                                                             gseaParam = 0.5)) +
                                     labs(title = paste0("GSEA of DEGs according to ", cfg$title_desc, " for ", cell_type, " cells using ", gene_set, " gene set."))
                                
                                # Define output directory and filename prefix
                                out_dir <- paste0(base_out_path, cfg$out_folder, "/", gene_set, "/")
                                file_prefix <- paste0(out_dir, gsub("/", "_", cell_type))
                                
                                ggsave(p, filename = paste0(file_prefix, ".pdf"), device = "pdf", width = 12, height = 8, create.dir = TRUE)
                                write_csv(sig_results, file = paste0(file_prefix, ".csv"))
                                
                                # clusterProfiler-style dotplot
                                plot_data <- sig_results[pathway %in% top_pathways]
                                
                                # Convert pathway to a factor ordered by NES
                                plot_data$pathway <- factor(plot_data$pathway, levels = plot_data$pathway[order(plot_data$NES)])
                                
                                p_cp_mimic <- ggplot(plot_data, aes(x = NES, y = pathway, color = padj, size = size)) +
                                        geom_point() +
                                        scale_color_gradient(low = "red", high = "blue", name = "p.adjust", guide = guide_colorbar(reverse = TRUE)) +
                                        scale_size_continuous(range = c(3, 8), name = "Set Size") +
                                        theme_minimal() +
                                        labs(title = paste0("GSEA of DEGs according to ", cfg$title_desc, "\nfor ", cell_type, " cells using ", gene_set, " gene set."),
                                             x = "Normalized Enrichment Score (NES)",
                                             y = NULL) +
                                        theme(axis.text.y = element_text(size = 10))
                                
                                ggsave(p_cp_mimic, filename = paste0(file_prefix, "_dotplot.pdf"), device = "pdf", width = 12, height = 8, create.dir = TRUE)
                        }
                }
        }
}
```


## ORA

```{r}
# Output path
base_ora_path <- paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/", annotation, "_annotation/enrichment/ora/")

analyses         <- c("sc", "pb", "both")
class_columns    <- list("sc"="class_sc", "pb"="class_pb", "both"="class_pb")
sig_columns      <- list("sc"="is_sig_sc", "pb"="is_sig_pb", "both"="both_sig")

# Get all valid human gene symbols from the database
valid_symbols <- AnnotationDbi::keys(org.Hs.eg.db, keytype = "SYMBOL")

# Prepare background universe set (all genes that ahve been tested in DE analysis)
universe_symbol <- rownames(data@assays$RNA@data)
universe_symbol <- intersect(universe_symbol, valid_symbols)

universe_entrez <- sig_entrez <- clusterProfiler::bitr(universe_symbol, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

for (cell_type in unique(total_DEGs$celltype)) {
        
        # Prepare Cytopus term2gene once per cell type
        cytopus_list <- c(cytopus_gene_sets[[cell_type]], cytopus_gene_sets[["global"]])
        cytopus_list <- lapply(cytopus_list, function(x) as.character(unlist(x, use.names = FALSE)))
        
        cytopus_term2gene <- data.frame(
                term = rep(names(cytopus_list), lengths(cytopus_list)),
                gene = unlist(cytopus_list, use.names = FALSE)
        )
        
        for (analysis in analyses) {
                
                # Loop through both UP and DOWN directions
                for (dir in c("UP", "DOWN")) {
                        
                        # Extract significant DEGs for the current analysis type and direction
                        sig_genes <- total_DEGs %>% 
                                dplyr::filter(.data[[sig_columns[[analysis]]]] == TRUE,
                                              .data[[class_columns[[analysis]]]] == dir,
                                              celltype == cell_type) %>% 
                                dplyr::pull(gene)
                        
                        # Keep only the genes that actually exist in the database
                        mapped_genes <- intersect(sig_genes, valid_symbols)
                        
                        # Skip to the next iteration if there are no significant genes for this set
                        if (length(mapped_genes) == 0) next 
                        
                        # Convert to ENTREZ ID for REACTOME
                        sig_entrez <- clusterProfiler::bitr(mapped_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
                        
                        # Initialize a list to store results for GO, REACTOME and CYTOPUS
                        ora_results <- list()
                        
                        # 1. Run ORA against GO
                        ora_results[["GO"]] <- enrichGO(gene      = mapped_genes,
                                                        OrgDb     = org.Hs.eg.db,
                                                        minGSSize = 5,
                                                        universe  = universe_symbol,
                                                        keyType   = "SYMBOL",
                                                        ont       = "BP")
                        
                        # 2. Run ORA against REACTOME
                        if (nrow(sig_entrez) > 0) {
                                ora_results[["REACTOME"]] <- enrichPathway(gene         = sig_entrez$ENTREZID,
                                                                           minGSSize    = 5,
                                                                           universe     = universe_entrez$ENTREZID,
                                                                           pvalueCutoff = 0.05,
                                                                           readable     = TRUE)
                        }
                        
                        # 3. Run ORA against Cytopus
                        ora_results[["CYTOPUS"]] <- enricher(mapped_genes,
                                                             minGSSize = 5,
                                                             universe  = universe_symbol,
                                                             TERM2GENE = cytopus_term2gene)
                        
                        # Iterate over the completed results to validate, plot, and save
                        for (db_name in names(ora_results)) {
                                res <- ora_results[[db_name]]
                                
                                # Check if the result object is valid and contains significant adjusted p-values
                                if (!is.null(res) && nrow(res@result) > 0 && min(res@result$p.adjust, na.rm = TRUE) < 0.05) {
                                        
                                        # Reduce redundancy specifically for GO
                                        if (db_name == "GO") {
                                                res <- clusterProfiler::simplify(res)
                                        }
                                        
                                        # Format the title text based on direction
                                        dir_text <- ifelse(dir == "UP", "up-regulated", "down-regulated")
                                        
                                        p <- dotplot(res, showCategory=20, font.size=10, label_format=70) +
                                                scale_size_continuous(range=c(1, 7)) +
                                                theme_minimal() +
                                                ggtitle(paste0(db_name, " Enrichment of ", dir_text, " genes for ", cell_type, " cells."))
                                        
                                        # Define output directory and filename prefix
                                        out_dir     <- paste0(base_ora_path, analysis, "/", db_name, "/")
                                        file_prefix <- paste0(out_dir, gsub("/", "_", cell_type), "_", dir)
                                        
                                        ggsave(p, filename = paste0(file_prefix, ".pdf"), device = "pdf", width = 12, height = 8, create.dir = TRUE)
                                        write_csv(res@result, file = paste0(file_prefix, ".csv"))
                                }
                        }
                }
        }
}
```






































