Filippo Gastaldello - 26/01/26

Analysis of the output of SPECTRA

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggpubr)
library(ggrepel)
```

# Data loading and preprocess
```{r}
# Cell scores quantify the activation of each gene program in every cell
cell_scores <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA/full/cytopus_annotation/Natri_et_al_full_cytopus_anno_cell_scores_15098.hpc3-head1.unitn.it.csv") %>% rename("cell_id"="...1")
# Expression metadata are needed to match cell id to their cell type
metadata <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/Natri_et_al_metadata.csv")
# Only keep metadata for cells in the SPECTRA output
metadata <- metadata %>% filter(cell_id %in% cell_scores$cell_id)
# Add condition column to cell scores
cell_scores <- cell_scores %>% 
                        left_join(metadata %>% select(cell_id, Diagnosis, Sample_Name, Phase, cytopus)) %>%
                        relocate(Diagnosis, .before = cell_id) %>%
                        relocate(Phase, .after = Diagnosis) %>%
                        relocate(Sample_Name, .before = Diagnosis) %>% 
                        relocate(cytopus, .after=Diagnosis)
# Remove programs without a name
unknown_programs <- grep("\\d+$", colnames(cell_scores)[6:length(colnames(cell_scores))], value = TRUE)
cell_scores <- cell_scores %>% select(-c(unknown_programs))
```
# Upnaired Welch's test

Test IGP activation differences in case vs control, for each cell type

```{r}
res_welch <- lapply(unique(cell_scores$cytopus),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% filter(cytopus == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "t-statistic"=numeric(),
                                            "df"=numeric(), "p"=numeric(), "control_avg"=numeric(), "case_avg"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Try to compute the test, it might fail if not enough cells are active
                                test_res <- try(t.test(x = (celltype_cell_score %>% filter(Diagnosis == "Control" & .[[program]]>0.001) %>% select(all_of(program))),
                                                       y = (celltype_cell_score %>% filter(Diagnosis == "IPF" & .[[program]]>0.001) %>% select(all_of(program))),
                                                       var.equal = FALSE),
                                                       silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "t-statistic"=NA, "df"=NA,
                                                                  "p"=NA, "control_avg"=NA, "case_avg"=NA)
                                                       )
                                }else{
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "t-statistic"=test_res$statistic, "df"=test_res$parameter,
                                                                  "p"=test_res$p.value, "control_avg"=test_res$estimate[1], "case_avg"=test_res$estimate[2])
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_welch <- bind_rows(res_welch)

# Compute FDR to account for multiple testing
IGP_activation_welch <- IGP_activation_welch %>%
                                drop_na() %>% 
                                group_by(cell_type) %>% 
                                mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                                ungroup() %>% 
        relocate(FDR, .after = p)

# Compute two measure of effect size
#
# logFC <- log(case_avg/control_avg)
# delta_active <- active_cells_case_perc - active_cells_control_perc
IGP_activation_welch <- IGP_activation_welch %>% mutate(delta_active = active_cells_case_perc - active_cells_control_perc,
                                                        logFC = log2(case_avg/control_avg))

write_csv(IGP_activation_welch, file = "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/IPG_activation_welch_test.csv")
```

## Boxplots - facet

Plot activation values in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
# Extract significant tests
welch_sig <- IGP_activation_welch %>% filter(FDR <0.05 & (abs(logFC) > 1 | abs(delta_active) > 20))
# Flatten cell_scores for plotting
plot_data <- cell_scores %>% pivot_longer(6:length(colnames(cell_scores)), names_to = "program")
# Only keep data points representing active cells ( activation > 0.001) belonging to significant comparisons
plot_data <- plot_data %>% inner_join(welch_sig %>% select(cell_type, program), by = join_by("cytopus"=="cell_type", "program"=="program")) %>% rename("cell_type"="cytopus") %>% filter(value>0.001)

p <- plot_data %>% ggplot(aes(x=program, y=value, fill=Diagnosis))+
                geom_boxplot()+
                theme(
                        axis.text.x = element_text(hjust=1, angle = 45, size = 5)
                )+
                facet_wrap(~cell_type, scale="free")
p
ggsave(filename = "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/welch_test/boxplots/boxplot_welch_significant_facet_celltype.pdf", plot = p, device = "pdf", width = 18, height = 15, create.dir = TRUE)
```

## Boxplots - per celltype

Plot activation values in case vs control, only in comparisons where the difference is significant according to the Welch test

```{r}
for (celltype in unique(metadata$cytopus)) {
        
        p <- plot_data %>% filter(cell_type==celltype) %>% ggplot(aes(x=program, y=value, fill=Diagnosis))+
                                                                geom_boxplot()+
                                                                labs(
                                                                        title = paste0("IGP with significant differences in activation between case and control in ",celltype," cells."),
                                                                        y = "Activation level"
                                                                )+
                                                                theme(
                                                                        axis.text.x = element_text(hjust=1, angle=45, size=8)
                                                                )+
                                                                stat_compare_means(
                                                                        aes(group = Diagnosis),
                                                                        method = "t.test",
                                                                        method.args = list(var.equal=FALSE),
                                                                        vjust = -1
                                                                )+
                                                                stat_summary(
                                                                    fun.data = function(y) {
                                                                      return(data.frame(
                                                                        y = max(y),  # Position: slightly above the max value of that group
                                                                        label = paste0("n= ",length(y)) # Label: the count of points
                                                                      ))
                                                                    },
                                                                    geom = "text",
                                                                    position = position_dodge(width = 0.75), # Must match your boxplot width (default is 0.75)
                                                                    vjust = -0.5,             # Adjust vertical spacing
                                                                    size = 3                  # Font size
                                                                  )
                                                                                        
        
        ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/welch_test/boxplots/boxplot_welch_significant_", celltype,".pdf"), plot = p, device = "pdf", width = 12, height = 8, create.dir = TRUE)
}
```
## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_welch %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & logFC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & logFC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=logFC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Welch's t-test",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & logFC>FC_thresh) | (FDR<FDR_thresh & logFC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/welch_test/volcano_welch.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```

# Linear Mixed Model

In Welch's test each cell is seen as an inpdendent measurement, ingoring the possible collinearity of cells belonging to the same patient. Linear Mixed Models correct for this modeling the patient id as a random effect.

```{r message=FALSE, warning=FALSE}
res_lmm <- lapply(unique(metadata$cytopus),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% filter(cytopus == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "term"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "avg_control_activation"=numeric(), "avg_case_activation"= numeric(), "estimate"=numeric(),
                                            "std.err"=numeric(), "statistic"=numeric(), "p"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Compute average activation in case vs control
                                avg_control_activation <- celltype_cell_score %>%
                                                                filter(Diagnosis == "Control") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                avg_case_activation <- celltype_cell_score %>%
                                                                filter(Diagnosis == "IPF") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                # Try to compute the test, it might fail if not enough cells are active
                                data <- celltype_cell_score %>% select(Diagnosis, Sample_Name, all_of(program)) %>% filter(.[[program]]>0.001)
                                colnames(data) <- c("Diagnosis", "Sample_Name", "Program")
                                test_res <- try(lmer(Program ~ Diagnosis + (1|Sample_Name), data = data),
                                                silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, term = NA, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "avg_control_activation"=avg_control_activation,
                                                                  "avg_case_activation"=avg_case_activation, "estimate"=NA, "std.err"=NA,"statistic"=NA, "p"=NA)
                                                       )
                                }else{
                                        test_res <- tidy(test_res, effects = "fixed")[2,]
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "term"=test_res$term,
                                                                  "active_cells_control_perc"=perc_active_cells_control, "active_cells_case_perc"=perc_active_cells_case,
                                                                  "avg_control_activation"=avg_control_activation,"avg_case_activation"=avg_case_activation,
                                                                  "estimate"=test_res$estimate, "std.err"=test_res$std.error, "statistic"=test_res$statistic, "p"=test_res$p.value)
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_lmm <- bind_rows(res_lmm)
IGP_activation_lmm <- IGP_activation_lmm %>% 
                        drop_na() %>% 
                        group_by(cell_type) %>% 
                        mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                        ungroup()
# Compute effect size as activationFC log2((avg_control_activation + estimate)/avg_case_activation)
IGP_activation_lmm <- IGP_activation_lmm %>% mutate(logFC = log2((avg_control_activation+estimate)/avg_control_activation))

write_csv(IGP_activation_lmm, file = "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/IPG_activation_lmm.csv")
```

## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_lmm %>%
                filter(!logFC=="NaN" ) %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & logFC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & logFC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=logFC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Linear Mixed Model - Sample ID as random effect",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & logFC>FC_thresh) | (FDR<FDR_thresh & logFC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/lmm/volcano_lmm.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```


# Linear Mixed Model - cell cycle phase correction

```{r message=FALSE, warning=FALSE}
res_lmm <- lapply(unique(metadata$cytopus),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% filter(cytopus == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "term"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "avg_control_activation"=numeric(), "avg_case_activation"= numeric(), "estimate"=numeric(),
                                            "std.err"=numeric(), "statistic"=numeric(), "p"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Compute average activation in case vs control
                                avg_control_activation <- celltype_cell_score %>%
                                                                filter(Diagnosis == "Control") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                avg_case_activation <- celltype_cell_score %>%
                                                                filter(Diagnosis == "IPF") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                # Try to compute the test, it might fail if not enough cells are active
                                data <- celltype_cell_score %>% select(Diagnosis, Sample_Name, Phase, all_of(program)) %>% filter(.[[program]]>0.001)
                                colnames(data) <- c("Diagnosis", "Sample_Name", "Phase", "Program")
                                test_res <- try(lmer(Program ~ Diagnosis + Phase + (1|Sample_Name), data = data),
                                                silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, term = NA, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "average_control_activation"=avg_control_activation,
                                                                  "avg_case_activation"=avg_case_activation, "estimate"=NA, "std.err"=NA, "statistic"=NA, "p"=NA)
                                                       )
                                }else{
                                        test_res <- tidy(test_res, effects = "fixed")[2:3,]
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "term"=test_res$term,
                                                                  "active_cells_control_perc"=perc_active_cells_control, "active_cells_case_perc"=perc_active_cells_case,
                                                                  "average_control_activation"=avg_control_activation, "avg_case_activation"=avg_case_activation,
                                                                  "estimate"=test_res$estimate, "std.err"=test_res$std.error, "statistic"=test_res$statistic, "p"=test_res$p.value)
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_lmm_phase_corrected <- bind_rows(res_lmm)
IGP_activation_lmm_phase_corrected <- IGP_activation_lmm_phase_corrected %>% 
                        drop_na() %>% 
                        group_by(cell_type) %>% 
                        mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                        ungroup()
# Compute effect size as activationFC log2((avg_control_activation + estimate)/avg_case_activation)
IGP_activation_lmm_phase_corrected <- IGP_activation_lmm_phase_corrected %>% mutate(logFC = log2((avg_control_activation+estimate)/avg_control_activation))

write_csv(IGP_activation_lmm_phase_corrected, file = "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/IPG_activation_lmm_phase_corrected.csv")
```

## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_lmm_phase_corrected %>%
                filter(!logFC=="NaN",
                       !str_detect(term, "Phase")) %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & logFC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & logFC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=logFC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed') +
                labs(
                        title = "Linear Mixed Model - Sample ID as random effect, correcting for cell cycle phase",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        ) +
                geom_label_repel(
                        data = plot_data %>% filter((FDR<FDR_thresh & logFC>FC_thresh) | (FDR<FDR_thresh & logFC< -FC_thresh)),
                        aes(label = tag),
                        size = 2.5,           
                        max.overlaps = Inf,   
                        box.padding = 0.5,
                        point.padding = 0.2,
                        min.segment.length = 0,
                        nudge_x = .15,
                        nudge_y = .5,) +
                theme_minimal()
p
ggsave(filename = paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/lmm/volcano_lmm_phase_corrected.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```





























