Filippo Gastaldello - 26/01/26

Analysis of the output of SPECTRA

```{r}
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggpubr)
library(ggrepel)
library(parallel)
library(Seurat)
library(eulerr)
library(patchwork)
library(ggplotify)
library(jsonlite)
library(tidyverse)
library(ggrastr)
library(msigdbr)
library(fgsea)
library(clusterProfiler)  
library(org.Hs.eg.db)
library(DOSE)
library(ReactomePA)
library(enrichplot)
library(effectsize)
```

```{r}
output_dir <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation/unknown_programs/"
```


# Data loading and preprocess
```{r}
# Define cell type annotation and gene sets to be used
annotation <- "manual"
annotation_column <- "manual_annotation_1"

# Cell scores quantify the activation of each gene program in every cell
cell_scores <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA/full/manual_annotation/Natri_et_al_full_manual_anno_cell_scores_93908.hpc3-head1.unitn.it.csv") %>% dplyr::rename("cell_id"="...1")
# Expression metadata are needed to match cell id to their cell type
metadata <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/Natri_et_al_metadata.csv")
# Only keep metadata for cells in the SPECTRA output
metadata <- metadata %>% filter(cell_id %in% cell_scores$cell_id)
# Add condition column to cell scores
cell_scores <- cell_scores %>% 
                        left_join(metadata %>% dplyr::select(cell_id, Diagnosis, Sample_Name, Phase, annotation_column)) %>%
                        relocate(Diagnosis, .before = cell_id) %>%
                        relocate(Phase, .after = Diagnosis) %>%
                        relocate(Sample_Name, .before = Diagnosis) %>% 
                        relocate(annotation_column, .after=Diagnosis)
colnames(cell_scores)[3] <- "annotation"
# only keep programs without a name
unknown_programs <- grep("\\d+$", colnames(cell_scores)[6:length(colnames(cell_scores))], value = TRUE)
cell_scores <- cell_scores %>% dplyr::select(all_of(c(colnames(cell_scores)[1:5], unknown_programs)))
```

# Name unknown programs

## Get factors markers
```{r}
markers <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA/full/manual_annotation/Natri_et_al_full_manual_anno_factor_markers_93908.hpc3-head1.unitn.it.csv") %>% dplyr::select(all_of(unknown_programs))
```


## ORA of factors markers

```{r}
cytopus_gene_sets <- read_json("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/cytopus_gene_sets/knowledgebase_gene_sets_manual_annotation.json")
universe_symbol <- markers %>% 
                        pivot_longer(everything()) %>% 
                        pull(value) %>% 
                        unique()

res <- lapply(unknown_programs, function(factor){
        
        cell_type <- str_split_i(factor, "-X-", 2)
        cell_type <- ifelse(cell_type == "global", "all", cell_type)
        
        # Prepare Cytopus term2gene for current cell type (program specificity)
        if(cell_type=="all") {
                cytopus_list <- c(cytopus_gene_sets[["global"]])
        }else{
                cytopus_list <- c(cytopus_gene_sets[[cell_type]], cytopus_gene_sets[["global"]])
        }
        cytopus_list <- lapply(cytopus_list, function(x) as.character(unlist(x, use.names = FALSE)))
        
        cytopus_term2gene <- data.frame(
                term = rep(names(cytopus_list), lengths(cytopus_list)),
                gene = unlist(cytopus_list, use.names = FALSE)
        )
        # Run ORA against Cytopus using factor markers
        ora_results <- enricher(markers %>% pull(factor),
                                minGSSize = 5,
                                universe  = universe_symbol,
                                TERM2GENE = cytopus_term2gene)
        
        if(is.null(ora_results)){ return(data.frame(fact = factor, name = factor)) }
        
        ora_results <- ora_results@result %>% 
                        filter(p.adjust <= 0.05) %>% 
                        mutate(GeneRatio = as.numeric(str_split_i(GeneRatio, "/",1))/as.numeric(str_split_i(GeneRatio, "/",2))) %>% 
                        slice_min(GeneRatio) %>% 
                        pull(ID)
        
        ora_results <- ifelse(is_empty(ora_results), str_split_i(factor, "-X-",1), ora_results)                
        
        return(data.frame(fact = factor,
                          name = paste0(str_split_i(factor, "-X-",1),
                                        "-X-",
                                        str_split_i(factor, "-X-",2),
                                        "-X-",
                                        ora_results)))
})

factor_names <- bind_rows(res)

colnames(cell_scores)[6:ncol(cell_scores)] <- factor_names$name
```

# Wilcoxon-Mann-Whitney test (DAP, sample aggregated)

Test IGP activation differences in case vs control for each cell type aggregating score at sample level

```{r}
res_WMW_sample_agg <- lapply(unique(cell_scores$annotation), function(cell_type){
        
        # Subset cell scores to keep only current cell type
        celltype_cell_score <- cell_scores %>% dplyr::filter(annotation == cell_type)
        
        # Store total cells for percentages
        n_total_control <- sum(celltype_cell_score$Diagnosis == "Control")
        n_total_ipf     <- sum(celltype_cell_score$Diagnosis == "IPF")
        
        tests_list <- list()
        
        programs <- colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]
        
        for (program in programs) {
        
                # Extract active cells
                vec_control <- celltype_cell_score %>% 
                                        dplyr::filter(Diagnosis == "Control" & .[[program]] > 0.001) %>%
                                        dplyr::select(all_of(program), Sample_Name)
                vec_ipf     <- celltype_cell_score %>%
                                        dplyr::filter(Diagnosis == "IPF" & .[[program]] > 0.001) %>%
                                        dplyr::select(all_of(program), Sample_Name)
                
                n_active_control <- nrow(vec_control)
                n_active_case    <- nrow(vec_ipf)
                
                perc_active_control <- round((n_active_control / n_total_control) * 100, 2)
                perc_active_case    <- round((n_active_case / n_total_ipf) * 100, 2)
                
                # Aggregate cell scores per sample
                vec_control_agg <- vec_control %>% summarise(.by = Sample_Name, score = mean(.data[[program]]))
                vec_ipf_agg     <- vec_ipf     %>% summarise(.by = Sample_Name, score = mean(.data[[program]]))
                
                # Extract the numeric vectors for the test
                scores_control <- vec_control_agg %>% pull(score)
                scores_ipf     <- vec_ipf_agg %>% pull(score)
                
                # Try to compute the test
                test_res <- try(wilcox.test(x          = scores_control,
                                            y          = scores_ipf,
                                            conf.int   = TRUE,
                                            paired     = FALSE),
                                silent = TRUE)
                
                if (inherits(test_res, "try-error") || length(scores_control) == 0 || length(scores_ipf) == 0) {
                        stat       <- NA
                        pval       <- NA
                        es         <- NA
                        es_ci_low  <- NA
                        es_ci_high <- NA
                } else {
                        stat <- test_res$statistic
                        pval <- test_res$p.value
                        
                        # Compute effect size and 95% CI on the sample-aggregated data
                        es_res <- effectsize::rank_biserial(x = scores_ipf, 
                                                            y = scores_control,  # The second group is used as reference by default 
                                                            ci = 0.95)
                        
                        es         <- es_res$r_rank_biserial
                        es_ci_low  <- es_res$CI_low
                        es_ci_high <- es_res$CI_high
                }
                
                # Store results
                tests_list[[program]] <- data.frame(
                                        cell_type                 = cell_type, 
                                        program                   = program, 
                                        active_cells_control_perc = perc_active_control,
                                        active_cells_case_perc    = perc_active_case, 
                                        n_active_control          = n_active_control, 
                                        n_active_case             = n_active_case,
                                        n_samples_control         = length(scores_control),
                                        n_samples_case            = length(scores_ipf),
                                        avg_activation_control    = mean(scores_control),
                                        avg_activation_case       = mean(scores_ipf),
                                        statistic                 = stat, 
                                        p                         = pval,
                                        effect_size               = es,
                                        effect_size_CI_lower      = es_ci_low,
                                        effect_size_CI_upper      = es_ci_high
                )
        }
        return(bind_rows(tests_list))
})

# Bind all cell types together
IGP_activation_WMW_sample_agg <- bind_rows(res_WMW_sample_agg)

# Compute FDR and log2FC (Removed manual effect size math)
IGP_activation_WMW_sample_agg <- IGP_activation_WMW_sample_agg %>%
  drop_na(p) %>% 
  group_by(cell_type) %>% 
  mutate(FDR = p.adjust(p, method = "fdr")) %>% 
  ungroup() %>% 
  relocate(FDR, .after = p) %>%
  mutate(log2FC = log2(avg_activation_case / avg_activation_control))
if(! dir.exists(paste0(output_dir, "wilcoxon_test_sample_aggregated/"))){dir.create(paste0(output_dir, "wilcoxon_test_sample_aggregated/"), recursive = TRUE)}
write_csv(IGP_activation_WMW_sample_agg, file = paste0(output_dir, "wilcoxon_test_sample_aggregated/IGP_activation_WMW_test_sample_aggregated.csv"))
```

## Boxplots - facet

Plot activation values in case vs control, only in comparisons where the difference is significant

```{r}
# Extract significant tests
WMW_sample_agg_sig <- IGP_activation_WMW_sample_agg %>% dplyr::filter(FDR <0.05 & (abs(effect_size) > 0.3))

plot_data <- cell_scores %>%
                pivot_longer(6:length(colnames(cell_scores)), names_to = "program") %>%                           # Flatten cell_scores for plotting
                filter(value > 0.001) %>%                                                                         # Only keep active cells
                group_by(Sample_Name, annotation, program, Diagnosis) %>% 
                summarize(value = mean(value)) %>%                                                                # Aggregate activation score per sample, for each cell type
                ungroup() %>% 
                inner_join(WMW_sample_agg_sig %>%                                                                 # Only keep significant de-regulations
                                   dplyr::select(cell_type, program, FDR),
                           by = join_by("annotation" == "cell_type", "program" == "program"))

p <- plot_data %>% ggplot(aes(x=program, y=value, fill=Diagnosis))+
                geom_boxplot() +
                theme_minimal() +
                theme(
                        axis.text.x = element_text(hjust=1, angle = 45, size = 5)
                ) +
                facet_wrap(~annotation, scale="free")

ggsave(filename = paste0(output_dir, "wilcoxon_test_sample_aggregated/boxplots_DAP/boxplot_wilcoxon_significant_facet_celltype.pdf"), plot = p, device = "pdf", width = 18, height = 15, create.dir = TRUE)
```
## Boxplot - per celltype

```{r}
for (celltype in unique(plot_data$annotation)) {
        
        p <- plot_data %>%
                dplyr::filter(annotation==celltype) %>%
                ggplot(aes(x=program, y=value, fill=Diagnosis)) +
                        geom_boxplot(
                                outliers = FALSE
                        ) +
                        geom_jitter(
                                position = position_jitterdodge(),
                                alpha    = 0.5,
                                size     = 0.7
                                ) +
                        labs(
                                title = paste0("IGP with significant differences in activation between case and control in ",celltype," cells."),
                                y = "Activation level"
                                ) +
                        stat_compare_means(
                                aes(group = Diagnosis),
                                method = "wilcox.test",
                                method.args = list(paired=FALSE),
                                vjust = -1,
                                label = "p.signif"
                                ) +
                        stat_summary(
                            fun.data = function(y) {
                              return(data.frame(
                                y = max(y),
                                label = paste0("n= ",length(y))
                              ))
                            },
                            geom = "text",
                            position = position_dodge(width = 0.75),
                            vjust = -0.5,
                            size = 3
                          ) +
                        theme_minimal() +
                        theme(
                                axis.text.x = element_text(hjust=1, angle=45, size=8)
                                )
        
        ggsave(filename = paste0(output_dir, "wilcoxon_test_sample_aggregated/boxplots_DAP/boxplot_wilcoxon_significant_", gsub("/", "_", celltype),".pdf"),
               plot = p,
               device = "pdf",
               width = 12,
               height = 8,
               create.dir = TRUE)
}
```

## Volcano plot FDR vs effect size
```{r}
plot_data <- IGP_activation_WMW_sample_agg %>% 
                mutate(label = paste0("(", cell_type, ") ", str_split_i(program, "-X-", 3)),
                       sig = case_when(
                               FDR <= 0.05 & effect_size >= 0.5                      ~ "up_high",
                               FDR <= 0.05 & effect_size >=0.3 & effect_size < 0.5   ~ "up_mid",
                               FDR <= 0.05 & effect_size <= -0.5                     ~ "down_high",
                               FDR <= 0.05 & effect_size <= -0.3 & effect_size> -0.5 ~ "down_mid",
                               TRUE                                                  ~ "no_sig"
                       )) %>%
                dplyr::select(label,
                              FDR,
                              effect_size,
                              sig)

p <- plot_data %>% ggplot(aes(x = effect_size, y = -log10(FDR), color = sig)) +
                        geom_point() +
                        scale_color_manual(
                                values = c("up_high" = "#f00514", "up_mid" = "#d19f9f", "down_high" = "#6e02e0", "down_mid" = "#b6a9c9", "no_sig" = "#d4d4d4"),
                                breaks = c("down_high", "down_mid", "up_high", "up_mid"),
                                labels = c("Over activated - High confidence", "Over activated - Moderate confidence","Repressed - High confidence", "Repressed - Moderate confidence")
                                )  +
                        geom_label_repel(
                                data = plot_data %>% filter(sig == "up_high" | sig == "down_high"),
                                aes(label = label),
                                size = 2.5,           
                                max.overlaps = Inf,   
                                box.padding = 0.5,
                                point.padding = 0.2,
                                min.segment.length = 0,
                                nudge_x = .15,
                                nudge_y = .5,
                                ) +
                        labs(
                                title    = "Deregulated programs activations according to Spectra scores",
                                subtitle = "Wilcoxon unpaired test on sample aggregated scores per cell type.",
                                color    = "Deregulation effect"
                                ) +
                        theme_minimal()
ggsave(plot = p,
       filename = paste0(output_dir, "wilcoxon_test_sample_aggregated/volcano_FDR_effect_size.pdf"),
       create.dir = TRUE,
       height = 9,
       width = 13,
       device = "pdf")
```

# Linear Mixed Model - cell cycle phase correction (DAP)

```{r message=FALSE, warning=FALSE}
res_lmm <- lapply(unique(cell_scores$annotation),
                function(cell_type){
                        # Subset cell scores to keep only current cell type
                        celltype_cell_score <- cell_scores %>% dplyr::filter(annotation == cell_type)
                        # Prepare dataset to store Welch's test results
                        tests <- data_frame("cell_type"=character(), "program"=character(), "term"=character(), "active_cells_control_perc"=numeric(), "active_cells_case_perc"=numeric(), "avg_control_activation"=numeric(), "avg_case_activation"= numeric(), "estimate"=numeric(),
                                            "std.err"=numeric(), "statistic"=numeric(), "p"=numeric())
                        # For each program, test for difference in activation between case and control, only in active cells (activation > 0.001) as per Authors suggestion
                        for (program in colnames(celltype_cell_score)[6:length(colnames(celltype_cell_score))]) {
                                # Compute number of active cells (activation > 0.001) in this cell type
                                perc_active_cells_control <- round((celltype_cell_score %>%
                                                                        dplyr::filter(Diagnosis == "Control" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                perc_active_cells_case <-    round((celltype_cell_score %>%
                                                                        dplyr::filter(Diagnosis == "IPF" & .[[program]]>0.001) %>%
                                                                        rownames() %>%
                                                                        length())/length(rownames(celltype_cell_score))*100,
                                                                   2)
                                # Compute average activation in case vs control
                                avg_control_activation <- celltype_cell_score %>%
                                                                dplyr::filter(Diagnosis == "Control") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                avg_case_activation <- celltype_cell_score %>%
                                                                dplyr::filter(Diagnosis == "IPF") %>% 
                                                                pull(any_of(program)) %>% 
                                                                mean()
                                # Try to compute the test, it might fail if not enough cells are active
                                data <- celltype_cell_score %>% dplyr::select(Diagnosis, Sample_Name, Phase, all_of(program)) %>% dplyr::filter(.[[program]]>0.001)
                                colnames(data) <- c("Diagnosis", "Sample_Name", "Phase", "Program")
                                test_res <- try(lmer(Program ~ Diagnosis + Phase + (1|Sample_Name), data = data),
                                                silent = TRUE)
                                if (inherits(test_res, "try-error")) {
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, term = NA, "active_cells_control_perc"=perc_active_cells_control,
                                                                  "active_cells_case_perc"=perc_active_cells_case, "avg_control_activation"=avg_control_activation,
                                                                  "avg_case_activation"=avg_case_activation, "estimate"=NA, "std.err"=NA, "statistic"=NA, "p"=NA)
                                                       )
                                }else{
                                        test_res <- tidy(test_res, effects = "fixed")[2:3,]
                                        tests <- rbind(tests,
                                                       data_frame("cell_type"=cell_type, "program"=program, "term"=test_res$term,
                                                                  "active_cells_control_perc"=perc_active_cells_control, "active_cells_case_perc"=perc_active_cells_case,
                                                                  "avg_control_activation"=avg_control_activation, "avg_case_activation"=avg_case_activation,
                                                                  "estimate"=test_res$estimate, "std.err"=test_res$std.error, "statistic"=test_res$statistic, "p"=test_res$p.value)
                                                       )
                                }
                        }
                        return(tests)
                }
        )
IGP_activation_lmm_phase_corrected <- bind_rows(res_lmm)
IGP_activation_lmm_phase_corrected <- IGP_activation_lmm_phase_corrected %>% 
                        drop_na() %>% 
                        group_by(cell_type) %>% 
                        mutate(FDR = p.adjust(p, method = "fdr")) %>% 
                        ungroup()
# Compute effect size as activationFC log2((avg_control_activation + estimate)/avg_case_activation)
IGP_activation_lmm_phase_corrected <- IGP_activation_lmm_phase_corrected %>% mutate(log2FC = log2((avg_control_activation+estimate)/avg_control_activation))

if(! dir.exists(paste0(output_dir, "lmm/"))){dir.create(paste0(output_dir, "lmm/"), recursive = TRUE)}
write_csv(IGP_activation_lmm_phase_corrected, file = paste0(output_dir, "lmm/IPG_activation_lmm_phase_corrected.csv"))
```

## Volcano plot
```{r}
# create tag for labels and deregulation column
FC_thresh <- 1
FDR_thresh <- 0.05

plot_data <- IGP_activation_lmm_phase_corrected %>%
                dplyr::filter(!log2FC=="NaN",
                       !str_detect(term, "Phase")) %>% 
                mutate(tag = paste0(cell_type, "-", str_split_i(program, "-X-",3)),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC>FC_thresh, "UP", "NO"),
                       diff_activated = ifelse(FDR<FDR_thresh & log2FC< -FC_thresh, "DOWN", diff_activated))

p <- plot_data %>% ggplot(aes(x=log2FC, y=-log10(FDR), color = diff_activated)) +
                geom_point(size = 1) +
                scale_color_manual(values = c("red", "grey", "purple"),
                                   labels = c("Downactivated", "Not significant", "Overactivated")) +
                geom_vline(xintercept = c(-1, 1), col = 'grey', linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = 'grey', linetype = 'dashed')  +
                        geom_label_repel(
                                data = plot_data %>% filter(diff_activated == "UP" | diff_activated == "DOWN"),
                                aes(label = tag),
                                size = 2.5,           
                                max.overlaps = Inf,   
                                box.padding = 0.5,
                                point.padding = 0.2,
                                min.segment.length = 0,
                                nudge_x = .15,
                                nudge_y = .5,
                                ) +
                labs(
                        title = "Linear Mixed Model - Sample ID as random effect, correcting for cell cycle phase",
                        subtitle = "comparison between activation in case vs control, for each program in each cell type",
                        x = "log2(activation FC)"
                        )+
                theme_minimal()

ggsave(filename = paste0(output_dir, "/lmm/volcano_lmm_phase_corrected.pdf"), plot = p, device = "pdf", width = 10, height = 10, create.dir = TRUE)
```
