---
author: Filippo Gastaldello
title: "lambda_variation"
output: html_document
---

```{r}
library(tidyverse)
```


Check results at different values of lambda (Spectra parameter)

# Paths and params
```{r}
# params
lambdas <- c("0.5", "0.1", "0.01", "0.001")

# paths
gigatables_paths <- paste0("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/manual_annotation_lam_",lambdas,"/gigatable.csv")
output_dir       <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/cytopus_genesets/lambda_analysis/"
```

# Read gigatables
```{r message=FALSE}
gigatables <- lapply(gigatables_paths, read_csv)
names(gigatables) <- lambdas
```

# Number of concordant programs

Check the number of concordant programs at different values of lambda. Programs are concordant if they are significant according to the Welch test with sample aggregated data (method from paper) and at least one of the GSEA analysis.
```{r}
res <- lapply(lambdas, function(lam){
        # Extract the gigatable for the current lambda
        gigatable <- gigatables[[lam]]
        # Count concordant programs
        concordant_programs_count <- gigatable %>%
                                        filter(effect_size_wmw_activation_sample_agg >= 0.3 & FDR_wmw_activation_sample_agg <= 0.05,
                                               if_any(c("padj_GSEA_sc", "padj_GSEA_pb"), ~ . <= 0.05)) %>% 
                                        nrow()
        return(data_frame("lambda" = lam, "count" = concordant_programs_count))
})
concordant_programs <- bind_rows(res)

p <- concordant_programs %>% ggplot(aes(x = lambda, y = count)) +
                                geom_bar(
                                        stat  = "identity",
                                        fill  = "orange",
                                        width = 0.4
                                        ) +
                                geom_text(
                                        aes(label = count),
                                        vjust = -1
                                        ) +
                                ylim(
                                        c(0, max(concordant_programs$count)+1)
                                        ) +
                                theme_minimal() +
                                theme(
                                        axis.text.y = element_blank()
                                        ) +
                                ggtitle(
                                        "Number of concordant programs between Spectra based and GSEA methods"
                                )
p
ggsave(filename = paste0(output_dir, "concordant_programs.pdf"),
       plot = p,
       width = 7,
       height = 4,
       create.dir = TRUE)
```

# Number of deregulated progragms

Plot the number of deregulated program according to different methods
```{r}
res <- lapply(lambdas, function(lam){
        gigatable <- gigatables[[lam]]
        return(data_frame("lambda" = lam,
                          "count"  = gigatable %>% filter(effect_size_wmw_activation_sample_agg>=0.3 & FDR_wmw_activation_sample_agg<=0.05) %>% nrow()
                          )
               )
})
deregulated_programs <- bind_rows(res)
p <- deregulated_programs %>% ggplot(aes(x = lambda, y = count)) +
                                geom_bar(
                                        stat  = "identity",
                                        fill  = "orange",
                                        width = 0.4
                                        ) +
                                geom_text(
                                        aes(label = count),
                                        vjust = -1
                                        ) +
                                ylim(
                                        c(0, max(deregulated_programs$count)+2)
                                        ) +
                                theme_minimal() +
                                theme(
                                        axis.text.y = element_blank()
                                        ) +
                                ggtitle(
                                        "Number of deregulated programs according to WMW sample aggregated"
                                )
p
ggsave(filename = paste0(output_dir, "deregulated_programs.pdf"),
       plot = p,
       width = 7,
       height = 4,
       create.dir = TRUE)
```

# Deregulation of critical programs at different values of lambda
```{r}
programs_of_interest <- c("fibro_TGFB1_response",
                          "M_mac_CSF1_response",
                          "fibro_EGF_response",
                          "lung-epi_TGFB1_response",
                          "all_MYC_targets",
                          "all_hypoxia-response",
                          "Mac_IL4-IL13_response",
                          "lung-epi_IL13_response",
                          "fibro_IL1B_response")

res <- lapply(lambdas, function(lam){
        
        gigatable <- gigatables[[lam]]
        # Which of the programs of interest are deregulated in WMW sample agg for this lambda?
        gigatable <- gigatable %>% 
                        filter(program_short %in% programs_of_interest,
                               FDR_wmw_activation_sample_agg<=0.05,
                               effect_size_wmw_activation_sample_agg>=0.3)
        
        df           <- data.frame(matrix(nrow = 0, ncol = 1+length(programs_of_interest)))
        colnames(df) <- c("lambda", programs_of_interest)
        df[1,1]      <- lam
        
        for (prog in unique(gigatable$program_short)) {
                df[1,prog] <- gigatable %>% filter(program_short==prog) %>% pull(cell_type) %>% paste(collapse = "\n")
        }
        return(df)
})
key_program_deregulation <- bind_rows(res)

p <- key_program_deregulation %>% 
        pivot_longer(cols = colnames(key_program_deregulation)[-1],
                     names_to = "program") %>% 
        rowwise() %>% 
        mutate(deregulations = length(str_split(value, ",")[[1]])) %>% 
        ungroup() %>% 
        mutate(deregulations = ifelse(is.na(value),0,deregulations)) %>% 
                ggplot(aes(x = lambda, y = program, fill = deregulations)) +
                        geom_tile() +
                        geom_text(
                                aes(label = value),
                                size = 3
                                ) +
                        scale_fill_gradient(
                                low   = "white",
                                high  = "orange",
                                guide = "none"
                                ) +
                        ggtitle(
                                "Program deregulation across different lambda values"
                                ) +
                        theme_minimal()
p
ggsave(filename = paste0(output_dir, "key_programs_deregulation.pdf"),
       plot = p,
       width = 14,
       height = 13,
       create.dir = TRUE)
```

