Filippo Gastaldello - 04/03/2026

Relative to SPECTRA run 93908

Analyis of gene program "lung-epi TGFB1 response" activation in Transitional AT2 cells and KRT5−/KRT17+ cells


```{r}
library(tidyverse)
library(eulerr)
library(ggVennDiagram)
library(ggplotify)
library(fgsea)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(msigdbr)

output_dir <- "/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/manual_annotation/IGP_interpretation/lung-epi_TGFB1_response/"
```

The gene program was identified twice in Transitional AT2 and once in KRT5-/KRT17+, respectively with index 419, 421 and 289 (This indexes may change by rerunning spectra). This most likely means that the three instances represent slightly different subprocesses of this gene program. Let's see how they differ by performing enrichment on the marker genes for each factor (as suggested in https://github.com/dpeerlab/spectra/issues/24 )

# Extract marker genes for each program
```{r}
IGP_of_interest <- c("419-X-Transitional AT2-X-lung-epi_TGFB1_response",
                     "421-X-Transitional AT2-X-lung-epi_TGFB1_response",
                     "289-X-KRT5-/KRT17+-X-lung-epi_TGFB1_response")

# Load marker genes
markers <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA/full/manual_annotation/Natri_et_al_full_manual_anno_factor_markers_93908.hpc3-head1.unitn.it.csv") %>% select(all_of(IGP_of_interest)) %>% as.list()
```
# Check overlap of marker genes
```{r}
# Compute overlaps
fit <- euler(markers)
# Plot venn
p <- as.ggplot(plot(fit, quantities = TRUE, fills = c("skyblue", "pink"), alpha = 0.5))
ggsave(plot = p, filename = paste0(output_dir, "markers_overlap.pdf"), device = "pdf", width = 8, height = 5, create.dir = TRUE)
```
# ORA

Let's perform Over Representation Analysis against GO and REACTOME for each set of markers.
```{r}
lapply(seq(1, length(markers)), function(i){
        factor_name    <- names(markers)[i]
        factor_markers <- markers[[i]] 
        
        # GO
        GO_results <- enrichGO(gene = factor_markers, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")
        p <- dotplot(GO_results, font.size=10, label_format=70) +
                scale_size_continuous(range=c(1, 7)) +
                theme_minimal() +
                ggtitle(paste0("GO Enrichment of factor markers for ",factor_name," cells."))
        ggsave(plot = p, filename = paste0(output_dir, "ORA/GO/", factor_name, ".pdf"), device = "pdf", height = 8, width = 12, create.dir = TRUE)
        
        # REACTOME (Need to convert symbol to EntrezID)
        factor_markers_entrez <- bitr(factor_markers, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
        REACTOME_results <- enrichPathway(gene = factor_markers_entrez$ENTREZID,
                                          pvalueCutoff = 0.05,
                                          pAdjustMethod = "BH",
                                          readable = T)
        p <- dotplot(REACTOME_results, font.size=10, label_format=70) +
                scale_size_continuous(range=c(1, 7)) +
                theme_minimal() +
                ggtitle(paste0("REACTOME Enrichment of factor markers for ",factor_name," cells."))
        ggsave(plot = p, filename = paste0(output_dir, "ORA/REACTOME/", factor_name, ".pdf"), device = "pdf", height = 8, width = 12, create.dir = TRUE)
})
```
# GSEA

Given that sets 419 and 421 are present in the same cell type (Transitional AT2) with opposite deregulation directions, let's merge the marker lists for the two factors, get DE analysis results for those genes in the cell type of interest, and run GSEA

Fetch GO and REACTOME gene sets
```{r}
# Fetch GO Biological Process 
go_bp_df <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
go_bp_list <- split(x = go_bp_df$gene_symbol, f = go_bp_df$gs_name)

# Fetch Reactome 
reactome_df <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")
reactome_list <- split(x = reactome_df$gene_symbol, f = reactome_df$gs_name)
```

```{r}
# Merge markers from factors 419 and 421
joined_markers <- c(markers[[1]], markers[[2]]) %>% unique()
# Get DE analysis results for these genes
DEGs <- read_csv("/shares/CIBIO-Storage/BCG/scratch/fgastaldello/COSBI/data/IPF_datasets/Natri_et_al/SPECTRA_results_analysis/full/manual_annotation/DE/sc_DEGs.csv")
DEGs <- DEGs %>% filter(celltype == "Transitional AT2") %>% filter(gene %in% joined_markers, FDR < 0.05)

# Prepare ranks
DEGs <- DEGs %>% arrange(desc(avg_log2FC))
rank <- DEGs$avg_log2FC
names(rank) <- DEGs$gene

# GO
set.seed(42)
fgsea_results <- fgsea(pathways = go_bp_list, 
                       stats    = rank,
                       minSize  = 15,
                       maxSize  = 500)
sig_results <- fgsea_results[padj < 0.05]
p <- as.ggplot(plotGseaTable(go_bp_list[sig_results$pathway],
                             rank,
                             fgsea_results,
                             gseaParam = 0.5)) +
                labs(title = paste0("GSEA of Transitional AT2's DEGs involved in TGFB1 response (GO:BP)"))
ggsave(plot = p, filename = paste0(output_dir, "GSEA/GO/Transitional_AT2_lung-epi_TGFB1_response.pdf"), device = "pdf", height = 8, width = 12, create.dir = TRUE)

# REACTOME
set.seed(42)
fgsea_results <- fgsea(pathways = reactome_list, 
                       stats    = rank,
                       minSize  = 15,
                       maxSize  = 500)
sig_results <- fgsea_results[padj < 0.05]
p <- as.ggplot(plotGseaTable(reactome_list[sig_results$pathway],
                             rank,
                             fgsea_results,
                             gseaParam = 0.5)) +
                labs(title = paste0("GSEA of Transitional AT2's DEGs involved in TGFB1 response (REACTOME"))
ggsave(plot = p, filename = paste0(output_dir, "GSEA/REACTOME/Transitional_AT2_lung-epi_TGFB1_response.pdf"), device = "pdf", height = 8, width = 12, create.dir = TRUE)
```

